Dual-color optical activation and suppression of neurons with high temporal precision

Version 1

Abstract

A fundamental challenge in the optogenetic toolbox is that all opsins, regardless of their excitation spectra, are activated by blue light. We hypothesized that pairing a red-shifted channelrhodopsin with a blue light-sensitive anion channel of appropriately matching kinetics shall render neurons responsive to a red but not blue light. To achieve this, we used a semi-rational mutagenesis strategy to optimize the kinetics and light spectrum of a chloride channelrhodopsin. By pairing optimized variants of blue light-sensitive anion channel ZipACR with vfChrimson, a fast red-shifted channelrhodopsin, we created a system in which red light derives precise and faithful action potentials of high frequencies, while blue light, through shunting inhibition, nullifies the effect of the red-shifted ChR2. Additionally, by a simple switch between red and blue lights, one can effectively excite or inhibit the activity of the same neurons.

Despite the short history of optogenetics, optical manipulation of neuronal activity has become an indispensable tool in the studies for functional analysis of brain circuits. The increasing needs for a more precise temporal and spatial manipulation of the neurons along with the untapped potentials of optogenetics have been driving forces in expanding the repertoire of optogenetics toolbox. Main development among these expansions is the discovery of opsins of diverse excitation spectra, with the current excitatory opsins showing light sensitivity ranging from near UV to near far-red spectra (Yizhar et al., 2011; Deisseroth and Hegemann, 2017; Marshel et al., 2019). The main purpose for such developments is to independently as well as collectively control the activity of distinct populations of neurons. This aim, however, has not been satisfied, mainly because all opsins, in addition to their own excitation spectral peak, are activated by blue light, a problem known in the field as cross-over or bleed-through. Attempts to create blue light-insensitive opsins will be most likely unsuccessful due to the intrinsic blue absorption by the retinal molecule, the non-encoded chromophore, which captures light energy in channelrhodopsins (Waddell et al., 1973; Wang et al., 2012).

There have been two stand out attempts to overcome the problem of cross-over. One uses a variant of blue-shifted ChR2 with fast kinetics in combination with a slow kinetics red-shifted ChR2 (R-ChR) (Klapoetke et al., 2014). The reasoning is that short pulses of blue light will activate only fast ChR2. Since it requires a tight regulation of the protein expression as well as the light intensity, this method has not been adopted for in-vivo application. Another approach is to expose axons expressing a R-ChR to prolonged photostimulation. Immediately following this, axons become unresponsive to light pulses and cease to fire. This method is limited to axonal regions as prolonged optical stimulation of the soma induces trains of action potentials (Hooks et al., 2015).

A potential solution to the problem of cross-over, the one that we propose here, is to co-express a blue-shifted anion channelrhodopsin (B-ACR) along with the R-ChR to ‘suppress’ its excitation by blue light. For this purpose, we chose vfChrimson-currently the fastest R-ChR-to gain more precise temporal control of red light excitation (Mager et al., 2018). This would, however, require a new variant of anion channelrhodopsin to optimize pairing. We used a semi-rational mutagenesis approach to modify the kinetic properties of ZipACR, an ultrafast blue light-activated chloride channel (Govorunova et al., 2017). We describe here two variants of ZipACR, I151T and I151V, with appropriate kinetics, light spectra, and photocurrent. We then paired the optimized anion channels with vfChrimson in a single expression unit. The new system, Zip-IvfChr, can drive time-locked high-frequency action potentials (APs) in response to red pulses (635 nm). In response to blue pulses (470 nm), on the other hand, Zip-IvfChr provides selective suppression of individual APs within trains of high frequency firing with high precision.

Results

The strategy to reduce blue light-mediated excitation by red-shifted channelrhodopsins

The channelrhodopsin Chrimson and its variants have the most red-shifted spectra (Klapoetke et al., 2014), and their response to 470 nm has a slower on-rate compared to the 590 nm light (Supplementary Fig. 1A-B). Despite this, 470 nm light is equally effective in inducing APs (Supplementary Fig. 1C). To overcome this limitation, we decided to take advantage of blue light-gated inhibitory chloride channels which effectively block action potential upon exposure to blue light (Govorunova et al., 2015; Govorunova et al., 2017). By co-expressing a B-ACR along with a Chrimson variant, photostimulation of targeted neurons by red light only activates R-ChR, which causes photoactivation of these neurons. Blue light, however, activates R-ChR as well as B-ACR. This allows the chloride channel to induce shunting inhibition of the neurons which in turn prevents firing of the neurons by blue light (Fig. 1A).

Reducing blue-light mediated excitation by red-shifted channelrhodopsins-Proof of concept.

(A) Schematic representation of the approach. A red-shifted channelrhodopsin (R-ChR) is co-expressed with a blue-shifted anion channel (B-ACR). When red light is ON, there is an overall excitation of the cell due to the more dominant R-ChR response. When blue-light is ON, the shunting inhibitory effect of B-ACR reduces the excitation induced by R-ChR to blue light. (B) Validation of the approach in flies expressing Chrimson, Chrimson and GtACR2, or GtACR2 alone . (B1) Experimental setup to record the courtship song of solitary male flies. (B2) Top, Example of a reconstructed neuronal arborization of P1 neurons, Bottom, example of a courtship song induced by LED pulses. (B3) Courtship song production of the solitary male flies during 10-sec of constant illumination with 590 nm (amber) or 470 nm (blue) light. Each dot represents an individual (mean ± SEM)

Validation of the co-expression strategy in Drosophila

As the first step for validating this approach, we took advantage of the well-studied male-specific P1 neurons in the fly central brain that elicit courtship singing upon activation (von Philipsborn et al., 2011; Inagaki et al., 2014; Ellendersen and von Philipsborn 2017; Mohammad et al., 2017) (Fig. 1B1-B2). When Chrimson was co-expressed with an inactive control protein (tdTomato), illumination at 470 nm and 590 nm wavelengths led to similar amounts of courtship song (respectively 110.1 ± 12.8 and 78.5 ± 11.6, n = 16-17), demonstrating that blue light at the chosen intensity was efficiently activating Chrimson (Fig 1B3). When Chrimson was co-expressed with the blue-sensitive anion channelrhodopsin GtACR2, 590 nm light illumination reliably elicited male courtship song (95.2 ± 9.14), however, 470 nm light illumination did not evoke any detectable pulse song (n = 17-18). Neither 590 nm nor 470 nm light pulses induced courtship song in flies expressing only GtACR2 (n = 5).

Identifying a suitable pair of ultrafast light-gated anion channel and a red-shifted Chrimson variant

Although GtACR2 proved to be effective in protocols with pulses of long duration, its slow current decay (Fig. 2A, see GtACR2, 161 ± 24 ms, n = 3) limits its application to optical stimulations of lower frequencies (Govorunova et al., 2015; Govorunova et al., 2017). For this reason, we searched for a pair of blue-shifted anion channelrhodopsin and red-shifted cation channelrhodopsin with fast and matching kinetics.

Comparison of channel properties of light-gated chloride channels and Chrimson variants.

(A) Representative photocurrents of the Chrimson and ACRs variants in whole-cell patch-clamp recordings from HEK293 cells with 590 nm and 470 nm illumination (10mW/mm2). For Chrimson variants, a re-conditioning pulse at 405 nm illumination was applied 15 seconds before 470 nm LED stimulation. The plot shows the off-time constant (top) following 1 sec of 470 nm illumination, and the peak photocurrent (bottom) at -60 mV (mean ± SD, n = 3-13 cells). (B) Typical response of CA1 cells from hippocampal organotypic slices expressing ZipACR to 40 Hz, 5 ms pulses (top black trace) and to an illumination protocol using 40 Hz of 10 ms light pulses at 470 nm (bottom blue trace). (C) Recordings from dentate gyrus granule cells in acute slices expressing ZipACR and vfChrimson. (C1) Representative responses to 500 ms illumination at 470 nm and 635 nm (blue and red traces respectively). Note, 635 nm illumination evoked action potential firing while 470 nm illumination only evoked a single action potential time-locked to the end of the light pulse (arrowhead). Insert, plot showing the number of light-evoked spikes at 635 nm (red bar) and 470 nm (blue bar) illumination. n = 11 cells, paired t-test **p< 0.01; (C2) Example of light-induced firings by 635 nm and 470 nm illumination (red and blue traces respectively, 5 ms pulses at 10 Hz). The arrowheads show consistent APs time-locked to the end of the blue light pulse. Insert, example of a voltage-clamp recording of currents evoked by 5 ms pulses of 635 nm and 470 nm light (red and blue traces respectively). Light-induced currents with 470 nm illumination were initially outward but turned inward immediately following the light offset. Note the late inward current matches the red light induced current, reflecting a slower kinetics for vfChrimson compare to ZipACR.

Characterization of the optimized variants of the ultrafast anion channelrhodopsin ZipACR.

(A) Structure-based design of ZipACR variants. (A1) The retinal binding pocket of GtACR1 and the putative retinal binding pocket of ZipACR homology model. (A2) The alignment of the transmembrane domains (TM) of GtACR1, GtACR2 and ZipACR. The residues surrounding the retinal binding pocket of GtACR1 and the corresponding residues from GtACR2 and ZipACR are highlighted in red. *marks the residues targeted for mutation. The lysine that forms the Schiff base with retinal is indicated in green. The numbers on the left correspond to the number of the first residue in each line. (B) Basic properties of the ZipACR variants compared to Ivf-Chrimson. off-time constant (top) following 1 sec of 470 nm illumination, peak photocurrent (middle) and 590/470 ratio (bottom). The data have been obtained from whole-cell recordings in HEK293 cells. Each dot corresponds to one cell (mean ± SD, n = 2-12).

(C) On-time constant (C1) and normalized photocurrent (C2) of the selected variants Zip(151T) and Zip(151V) as compared to the original ZipACR and IvfChrimson, with 470 nm or 635 nm illumination of various light intensities (in mW/mm2). For the photocurrent, the values have been measured at steady-state. The data points are the mean value ± SEM (n = 6-7). (D) The responses of the selected variants to various wavelengths of the same photon flux. (D1) The action spectra of ZipACR, Zip(151V) and Zip(151T) as measured from 650 nm to 400 nm (D2) For the action spectra, the maximum photocurrent amplitudes measured at each wavelength were normalized to the peak values obtained from the same cell across the spectrum. The data points are the mean value ± SEM (n = 6-7). (E) Validation of the efficiency of the Zip(151T) variant in slices. (E1) The response of dentate gyrus granule cells expressing Zip(151T) variant at 470 nm and 635 nm light pulses delivered through a 63x objective. (E2) Representative traces of the firing induced by current injection at 40 Hz. Overlapping 470 nm but not 635 nm pulses block action potentials. The plot on the right shows the efficiency (in %) of Zip(151T) in blocking individual action potentials at 470 and 635 nm light pulses (mean ± SD, n = 6-10 cells). The traces of the two outliers are presented in supplementary Fig. 5A.

For a fast R-ChR, we evaluated the recently published vfChrimson, the fastest red-shifted channelrhodopsin available at this time, which can induce APs up to the intrinsic limit of the neurons (Mager et al., 2018). Previously, we found that Chrimson and its variants traffic poorly to the membrane and developed a strategy to improve the trafficking (C-Chrimson-ts, Buonaventura et al., 2019). We adapted this modification to improve the trafficking of vfChrimson (referred as IvfChrimson). In our measurements in HEK293 cells, IvfChrimson has a channel off-rate consistent with previously reported (Mager et al., 2018) (off-rate time constant of 5.6 ± 0.3 ms, n = 10 compared to C-Chrimson-ts 21.5 ± 1.2, n = 10; mean ± SEM, Fig. 2A, Supplementary Fig. 2). When comparing C-Chrimson-ts and IvfChrimson photocurrents, we observe 1.68 fold reduction of normalized photocurrent of IvfChrimson (0.086 ± 0.055 pA/pF/A.U, n = 25) compared to C-Chrimson-ts (0.145 ± 0.126 pA/pF/A.U., n = 25; mean ± SD, Supplementary Fig. 2B) when the peak photocurrent amplitude was adjusted to membrane capacitance and membrane fluorescence.

To counteract the IvfChrimson excitation, we estimated that desired properties of the blue light activated anion ChR should have the off-rate time constant of ∼15-30 ms, minimal response to light of wavelength > 590 nm and a strong photocurrent. We identified ZipACR, RsACR_995, GcACR_457, RlACR_477 and GtACR2 as possible templates for further development based on published data (Govorunova et al., 2017). GtACR2, RsACR_995 and RIACR_477 are spectrally ideal with negligible response to 590 nm light (Supplementary Fig. 3A), but their kinetics are slow (mean off-rate of 161.0 ± 24.1, 122.7 ± 36.3 and 95.8 ± 49.4 ms, respectively, n = 3, Fig. 2A) which could have undesired long lasting inhibitory effects after light termination. GcACR_457 has very small photocurrents and cannot be characterized accurately, possibly due to poor membrane trafficking. ZipACR, although spectrally not ideal (∼515 nm; Govorunova et al., 2017) has a kinetic close to the desired range (9.5 ± 3.0 ms, n = 9) and it has strong photocurrents (Fig. 2A).

To test the inhibitory property of ZipACR, we injected 40 Hz somatic current in organotypic hippocampal neurons expressing the channel. Consistent with the original report, individual pulses of 470 nm light resulted in the time-locked suppression of APs (Fig. 2B).

Next we examined whether ZipACR can block blue light-mediated APs in neurons expressing vfChrimson. Recordings from the granule cells in acute hippocampal slices co-expressing ZipACR and vfChrimson showed that, compared with 635 nm light, 470 nm light induced significantly fewer APs (Fig. 2C1), in line with our prediction. However, we consistently observed the appearance of individual APs time-locked to the blue light offset (Fig. 2C1-C2). We concluded that due to the fast closing rate, the photoinhibitory current generated by ZipACR decays too early to suppress the remaining excitatory response of vfChrimson. Therefore, we decided to generate variants of ZipACR with slightly slower off-kinetics.

Optimizing ZipACR off-kinetics

It is technically easier to generate mutations with slower kinetics and blue-shifted spectra than generating variants with faster on-and off-kinetics. Thus, we decided to use ZipACR as a template for further mutagenesis instead of further modifying GtACR2. To identify relevant mutation sites in ZipACR, we generated a homology model of ZipACR with the crystal structure of GtACR1 (Govorunova et al., 2017) and identified the putative residues forming the retinal binding pocket in ZipACR (Fig. 3A). Multiple alignments were performed with the kinetically slower GtACR1 and GtACR2 to identify possible replacement of these residues that would result in slower kinetics and/or small blue-shift in ZipACR without disrupting the channel function (Fig. 3A2). We tested these variants in HEK293 cells for photocurrent ratio 590 nm / 470 nm LED stimulating light and measured the channel off-rate kinetics and photocurrent amplitude (Fig. 3B). Of these candidates, we identified I151V and I151T at the 4th transmem-brane domain of ZipACR that resulted in a small increase in channel off-rate time constant (15.5 ± 8.4 ms for I151V and 20.2 ± 6.4 ms for I151T, mean ± SD, Fig. 3B) compared to the original ZipACR (9.5 ± 2.9 ms, mean ± SD). We also generated double and triple mutants combining the I151V or I151T with Y170F, Y177F or V229L mutations but most of these double or triple mutants appeared to have smaller photocurrents and off-rate kinetics that could not be fitted well with single exponential back to zero baseline. Based on these results, we concluded the variants I151V (Zip(151V)) and I151T (Zip(151T)) are the best candidates for the pairing and proceeded for further characterizations.

Neither Zip(151V) nor Zip(151T) mutations changed the ion permeability of ZipACR as measured by channel reversal potential in potassium gluconate and cesium chloride based intracellular recording solutions (Supplementary Fig. 3B-C). In terms of light sensitivity, it is desirable that the ZipACR variant would respond to 470 nm light faster and stronger and to red-shifted light slower and weaker than IvfChrimson at the same intensity (Supplementary Fig. 4). With both Zip(151V) and Zip(151T), the channel on-rate kinetics have the desirable properties as described above to 470 nm and 590 nm light, although the Zip(151T) on-rate starts to approach the on-rate of IvfChrimson at light intensities >5 mW/mm2 at 590 nm (Fig. 3C1). With 470 nm light, IvfChrimson does not appear to reach saturating response even at 20 mW/mm2 (Fig. 3C2). With 590 nm light, IvfChrimson reaches saturating response at 10 mW/ mm2 whereas none of the ZipACR variants appears to reach saturating response even at 20 mW/mm2 (Fig. 3C2). Spectrally, Zip(151T) shows a wider red-shifted action spectra than ZipACR whereas Zip(151V) may have a slightly narrower blue-shifted action spectra than ZipACR (Fig. 3D).

Since Zip(151T) is a more red-shifted variant of the two (Fig. 3B,D), we asked whether red light pulses may induce photoinhibitory currents that suppress APs. We expressed Zip(151T) in the granule cells of the dentate gyrus in hippocampal slices using AAV vector and tested its inhibitory effect upon 470 nm and 635 nm light pulses. We injected somatic current at 40 Hz, while alternately delivering overlapping 5 ms pulses of light. Except in two instances, blue light pulses were consistently effective in blocking APs, while red light had no effect on APs induction (Fig. 3E, n = 10 for 470 and n = 6 for 635, the two outliers are presented in Supplementary Fig. 5A). It has been reported that at resting or hyperpolarized membrane potentials ZipACR may produce light-evoked APs (Kato et al., 2018). We did not observe such a phenomenon under our experimental setup (Supplementary Fig. 5B).

Validation of the co-expression strategy in brain slice

Since co-injection of two viruses results in variability in the ratio of expression in different neurons, we created constructs containing IvfChrimson and Zip(151T/V) (ZipT-IvfChr and ZipV-Ivf-Chr) using the bicistronic 2A sequence (Fig. 4A). This ensures an equimolar expression of the two channels. We also introduced a soma targeting Kv2.1 sequence into the constructs while eliminating the endoplasmic reticulum-exporting sequence of Ivf-Chrimson construct. Based on the photocurrent amplitudes of the constructs in response to 590 nm light (at -40mV holding potential with potassium gluconate-based intracellular solution to minimize Zip(151V) variant responses), the photocurrent amplitudes of IvfChrimson-only and ZipV-IvfChr constructs are comparable after adjustment to cell size and membrane fluorescence in HEK293 cells (Supplementary Fig. 6). We did not test the ZipT-IvfChr variant as the red-shifted response of Zip(151T) is difficult to eliminate to 590 nm light.

ZipACR variants co-expressed with the red ChR2 IvfChrimson prevent blue-light induced action potentials.

(A) The expression cassette for vfChrimson and bicistronic Zip-IvfChr variants used in rAAV vector (Top). The selected ZipACR variants (Zip(151T) or Zip(151V)) are co-expressed with the red-sensitive IvfChrimson by using the 2A self-cleaving peptide (T2A). All the opsins are soma-targeted with a Kv2.1 peptide. (Bottom) Representatives showing the membrane enriched expression of ZipT-IvfChr and ZipV-IvfChr. Scale bar = 25um. (B) Blue and red light-driven spike fidelity at various light intensities in DG cells (n = 4-10 cells for each opsin; 10 ms pulse width, single pulse). (C) Comparison of the high-frequency red and blue light-induced spiking (C1), and spike probability (C2, n = 6-13 for each opsin), of DG cells expressing vfChrimson or the Zip-IvfChr variants, with 10 and 20 Hz of light pulses (4-7mW/mm2). Holding potential = -60mV. (D) Optical inhibition of spiking in DG cells expressing the Zip-IvfChr variants. Representative traces of the firing induced by current injection at 40 Hz. Overlapping 470 nm but not 635 nm light pulses block action potentials. Holding potential = -60mV. The plots on the right show the efficiency (in %) of ZipT-IvfChr and ZipV-IvfChr in blocking individual action potentials at 470 and 635 nm (n = 9-10 for each opsin).

As the first step to test the performance of the system, we examined the power at which 470 nm and 635 nm lights induce APs in ZipT-IvfChr and ZipV-IvfChr, as well as in vfChrimson expressing neurons. We expressed AAV constructs in the granule cells of the dentate gyrus in the hippocampal slices (as described in Fig. 3E1). We recorded cells under the current-clamp mode and injected current through the patch pipette to maintain the membrane potentials at −60 mV. As expected, in vfChrimson expressing neurons 470 nm and 635 nm lights induced APs with light powers as low as 0,5mW/mm2 (n = 10, Fig. 4B). In contrast, ZipT-IvfChr expressing neurons produced APs only to 635 nm light stimulation, whereas 470 nm light powers as high as 10 mW/mm2 failed to induce APs (n = 7-8). As for ZipV-IvfChr, we occasionally observed blue light-induced APs (n = 4-7, Fig. 4B), which may be explained by its relatively faster decay time (Fig. 3B). In neurons expressing vfChrimson, we noticed a reduction in probability of AP upon successive optical stimulation specifically to 635 nm pulses (Fig. 4B). Consistent with this, we observed a reduction in response in HEK293 cells expressing IvfChrimson (Supplementary Fig. 7A,B). This phenomenon, likely caused by the desensitization of the channel, is consistent with previous reports (Lin et al., 2013).

To examine the efficiency of 635 nm and 470 nm lights in inducing high frequency APs, we delivered 10 Hz and 20 Hz light pulses of 5 ms duration (Fig. 4C). As expected, in vfChrimson expressing neurons both 470 nm and 635 nm lights induced APs (n = 8-9). In neurons expressing either ZipT-IvfChr or ZipV-IvfChr, 635 nm pulses were equally effective in driving APs. On the other hand, 470 nm light only induced modest depolarization, which never reached the threshold for AP induction (n = 5-12 for each opsin) (Fig. 4C2).

To test the temporal precision of the system in blocking AP, we stimulated the neurons by somatic current injection at 40 Hz, while alternately delivering perfectly overlapping 5 ms pulses of lights. Light pulses of 470 nm were consistently effective in blocking APs, while 635 nm light pulses had virtually no effect on Aps induction as its amplitude, half-width and delayed time were in-distinguishable from APs induced in the absence of light pulse (Fig. 4D).

Local mono-synaptic connection is a common feature in many brain regions. For this reason, independent dual optical activation requires that while optically stimulating pre-synaptic neurons, post-synaptic neurons remain responsive to the pre-synaptic inputs. To test whether our system meets this criterion, we asked how fast a neuron recovers from the inhibition induced by the ZipACR variants. We induced APs through somatic current injection at different time points from the offset of a light pulse. Blue light pulses blocked APs if the current was injected simultaneously. As expected from the off-rate of Zip(151T), within 5 ms from the light offset neurons expressing Zip(151T) or ZipT-IvfChr produced APs with high probability (Supplementary Fig. 8). The success rate reached its maximum within 10-15 ms. The recovery time for ZipV-IvfChr was immediate, with 100% success rate immediately following the light offset. This indicates that ZipV-Ivf-Chr could be used for dual optical activation without off target inhibition.

Discussion

The main purpose of this work was to find a practical solution to the problem of cross-over where all opsins, regardless of their preferred excitation spectra, are activated by blue light. We describe a R-ChR2 expression system by which pulses of red light drive time-locked high frequency APs, while blue light remains ineffective. We created this system by pairing a B-ACR with the ultrafast red-shifted ChR2, IvfChrimson. This requires the B-ACR to have a significantly larger photocurrent compared to R-ChR2 as well as a faster channel opening and slightly slower channel closing. For this reason, we chose ZipACR as the template B-ACR for further modifications. Based on our homology modelling and multiple alignments with the kinetically slower B-ACRs, GtACR1 and GTACR2, we identified potential residue replacements within the putative retinal binding pocket which would result in slightly slower kinetics without impairing the channel function.

To our surprise, many mutations of ZipACR around all-trans retinal do not alter the channel properties or compromise channel conductance too strongly. This is contrary to our past experiences with channelrhodopsins where conductance is highly sensitive to mutations around the retinal binding pocket (data not shown). ZipACR double mutants appear to have more compromised conductance or altered properties. Among the tested positions, the I151 residue appears to have significant influences in kinetics and spectral properties. Indeed, we found that two variants, Zip(151T) and Zip(151V), sit in a sweet spot for our purpose. More specifically, the channels off-rate kinetics have the desirable property of being slightly but sufficiently slower than Ivf-Chrimson without compromising the opening rate, photocurrent and light spectrum. Thus, when paired with IvfChrimson, they prevent AP induction by blue light, without interfering with red light-induced high frequency neural spiking. On a similar note, although blue light pulses produced subthreshold depolarization, we did not observe charge integration that could have resulted in AP, even in frequencies as high as 20 Hz; this most likely reflects the reverse potential of the ZipACR. Of the two variants, we observed Zip(151T), and consequently ZipT-IvfChr, to be more consistent in blocking blue light-induced APs, which could be explained by its slightly slower closing rate. Although Zip(151T) is the more red-shifted of the two, within the tested range of light intensities and expression, only rarely did we observe a block of APs at 635 nm. However, we do not recommend the use of orange light pulses, as we observed a significant photocurrent in this wavelength.

Based on its impact on the kinetics and spectral properties, it will be of value to conduct saturating mutagenesis study on I151 residue that may produce variants of interesting properties or pairing that could outperform Zip(151V) and Zip(151T).

A recent study (Vierock et al., 2020) proposed a R-ChR2 expression system by relying on the pairing between GtACR2 and Chrimson, as we did in our early stage work with Drosophila. For physiological as well as experimental reasons, the priority we set was to create a system with fast closing kinetics. As channelrhodopsins are known to conduct calcium ions, long opening lifetime can evoke significant elevation of intracellular calcium (Lin et al., 2009) and activate second messenger pathways with complicating biological effects in experiments of long duration. The same concern applies to B-ACRs with slow closing kinetics as they could have additional inhibitory effects after the termination of blue light and alter the physiology to be studied. From an experimental viewpoint, a fast photostimulation system has the flexibility to perform a wide range of optical manipulations. This includes driving neuronal activity to their firing limits as well as suppressing their firing with single spike precision. Al-though not tested here, our results indicate that the fast closing kinetics of the B-ACR channels, particularly ZipV-IvfChr, should allow dual optical activation of monosynaptically connected neurons.

Before concluding, we must note that our approach, which runs based on chloride influx is unlikely to be suitable for axonal terminal activation, where the chloride concentration is high (Mahn et al., 2016; Mahn et al., 2018). For this matter, we refer the readers to some published methods by which independent dual channel photostimulation of the axons might be feasible (Klapoetke et al., 2014; Hook et al., 2015). Thus, our system complements these studies as they are unsuitable for somatic photoactivation. This provides researchers with complementary methods to choose from based on their experimental needs.

Methods

Drosophila optogenetic activation experiments

Flies (Drosophila melanogaster) were kept on standard food medium (water, cornmeal, oatmeal, sucrose, yeast, agar, acetic acid, preservative methyl-4-hydroxybenzoate) at 25°C, 60% humidity, and a 12h/12h light-dark cycle. Male flies were collected after eclosion and aged for 7 days in isolation and protected from light on standard food, followed by 3 days on food containing 400 µM all-trans retinal (Sigma). The following genotypes were assayed: co-expression of Chrimson and GtACR2: w-; NP2631.GAL4/ UAS-CsChrimson.mVenus.attp40; tubP-FRT-stop-FRT-GAL80, fruFLP/ UAS-GtACR2.attp2; co-expression of Chrimson and td-Tomato: w-; NP2631.GAL4/ UAS-CsChrimson.mVenus.attp40; tubP-FRT-stop-FRT-GAL80, fruFLP/ UAS-myr-tdTomato. attp2; co-expression of GtACR2 and tdTomato: w-; NP2631. GAL4/ UAS-myr-tdTomato.attp2; tubP-FRT-stop-FRT-GAL80, fruFLP/ UAS-GtACR2.attp2.

For optogenetic activation experiments, we used a previously described multi-channel array of electret condenser microphones (CMP-5247TF-K, CUI Inc), amplified with a custom-made circuit board and digitized with a multifunction data acquisition device (NI USB-6259 MASS Term, National Instruments) (von Philipsborn et al., 2014). Single male flies were placed in recording chambers equipped with microphones at the bottom and light diodes (Amber 591 nm: Cree 5-mm C503B-AAS/AAN-015 or Blue 470 nm: Cree 5-mm C503B-BAS-CY0C0461) were placed 1 cm above the song recording chambers at a 60° angle, illuminating the whole chamber (10-sec of constant light). Tentative pulse song was detected by a MATLAB script and corrected manually, using a custom made user interface for visualization and annotation of song oscillograms (von Philipsborn et al., 2014).

Cell culture, transfection and stable cell line generation

HEK293 cells (Invitrogen) were grown in DMEM (1g/L glucose) supplemented with 8% FBS and 1% Pen/Strep. For recordings, cells were plated at low density on glass coverslips and kept in DMEM media supplemented with 4% FBS and 1% Pen/Strep. Transfection was done with Xtremegene 9 (Sigma-Aldrich) according to the manufacturer’s instruction. Recordings of the cells were performed between 38-50 hours after transfection.

For generation of recombinant lentivirus for stable HEK293 cell lines expressing IvfChrimson and ChrimsonR-tdTomato cells, HEK293 cells were co-transfected with psPAX2 and pMD2.G (courtesy of Trono laboratory) and the media was transferred to a new dish with untreated cells 2 and 3 days after transfection. The cells receiving the media were kept as stable cell lines of respective constructs.

Structure alignment and modelling

Homology modelling of ZipACR structure was performed with Modeller within the UCSF Chimera to the published GtACR1 structure (PDB #6CSM). This was further aligned to the GtACR2 alignment from Govorunova et al., (2017). Residues surrounding the retinal binding pocket of GtACR1 and the modelled ZipACR were identified and those residues that were different in ZipACR were changed to the corresponding residues of GtACR2 or GtACR1.

Molecular biology

For GcACR_457, RsACR_995 and RlACR_477, the cDNA were in the original pcDNA3.1 vector as EYFP fusion as available on Addgene (#103137, 103772, 103771). The GtACR2 construct was synthesized (IDT DNA) and modified with a N-terminal signalling peptide from the commercial pDisplay vector (Invitrogen) and a trafficking signal prior to tdTomato or citrine fusion and inserted in the pcDNA3.1+ vector (Invitrogen) for expression. For the ZipACR constructs, the original EYFP fusion from Addgene (#88844) was retained and the construct was either inserted into the pcDNA3.1+ vector or a second generation lentiviral vector pLenti with CMV promoter. No functional or fluorescent difference in the expressed protein were observed between expression from pcDNA3.1+ and pLenti vector in HEK293 cells and the results were combined for comparison and analysis.

Site-directed mutagenesis of the ZipACR, Chrimson and GtACR2 were performed with overlap extension PCR using Phusion DNA Polymerase (Life Technologies) and ligated into the pcDNA3.1+ or pLenti expression vector.

For ChrimsonR and vfChrimson, additional improvements for membrane trafficking was used as previously described (Buonaventura et al., 2019). In brief, N-terminal leading sequence from ChR1/oChIEF/ReaChR was transplanted to the transmembrane A-G of ChrimsonR and vfChrimson. Trafficking sequence was added after the transmembrane G before fusion to tdTomato. The modified ChrimsonR/vfChrimson-tdTomato constructs were inserted into the 2nd generation lentiviral pLenti vector with CMV promoter and used to make stable ChrimsonR/vfChrimson-td-Tomato HEK293 cell lines. The improved vf-Chrimson is named IvfChrimson.

For the comparison of IvfChrimson-citrine and Zip(151V)-Kv2.1-2A-IvfChrimson-citrine-Kv2.1 (ZipV-IvfChr) photocurrents, the two cDNAs were inserted into pcDNA3.1+ for expression. The IvfChrimson in both designs contains the modified N-terminal sequence from ChR1/oChIEF/ReaChR but the trafficking sequence was only used in the IvfChrimson-citrine design and not the Kv2.1 design.

rAAV production

rAAV2-retro containing vfChrimson_EYFP_Kv2.1, Zip(151T)-Kv2.1_IRES_DsRed, Zip(151T)-Kv2.1-2A-IvChrimson-citrine-Kv2.1 (ZipT-IvfChr), and Zip(151V)-Kv2.1-2A-IvfChrimson-citrine-Kv2.1 (ZipV-IvfChr) were produced according to the protocol at https://www.salk.edu/science/core-facilities/viral-vector-core/resources/. In brief, HEK293A cells (Life Technologies, Carlsbad, CA) were grown to 90% confluency and transfected with vectors containing opsin proteins, and the helper plasmids XX6-80 and rAAV2-retro (gift from A. Y. Karpova, Howard Hughes Medical Institute Janelia Farm Research Campus). rAAV2-retro were released and purified according to protocol at https://www.takarabio.com/documents/User%20Manual/6232/6232_e.v1509Da.pdf.

Electrophysiological characterization and imaging of HEK293 cells

Whole-cell patch-clamping was performed with Axopatch 200A patch-clamp amplifier with Digidata 1322A data acquisition board and pCLAMP10 software. Cells were clamped at -60mV except in experiments with the ramp protocol and experiments comparing the 590 nm light-induced photocurrents between Ivf-Chrimson-citrine and ZipV-IvfChr expressing cells. In the experiment comparing the IvfChrimson-citrine photocurrent to the 2A configuration, cells were clamped at -40mV to reduce the influence of ZipACR response with IvfChrimson photocurrent. Series resistance for the recorded cells were mostly under 15 MΩ and series resistance were compensated up to 70% during recordings where possible.

The extracellular solution was composed of (in mM): 145 NaCl, 3 KCl, 2 CaCl2, 1 MgCl2, 20 D-glucose and 10 HEPES (pH 7.35). For most experiments, the intracellular solution contained (in mM): 110 CsCl, 15 TEACl, 10 HEPES, 5 Cs2EGTA, 1 MgCl2 and 0.1mM CaCl2 (pH 7.25). In experiments comparing the 590 nm light-induced photocurrents between IvfChrimson-citrine and ZipV-IvfChr expressing cells and some ramp experiments, a K-gluconate based intracellular solution containing: 115 K-gluconate, 10 K2EGTA, 10 HEPES, 5 NaCl, 10 KCl, 2 MgATP and 0.1 Tris-GTP (pH 7.25), was used.

Channelrhodopsin-expressing cells were imaged and recorded on an Olympus IX73 microscope with a 40x air objective equipped with Photometrics Cascade II 512 EMCCD camera. Images were acquired using Micromanager 1.4 with no gain and no electron multiplication and exposure time of 250 ms. Excitation light was provided by a Thorlabs 4 channel LED (LED4D118 with 405, 470, 530 and 590 nm channels and only 470 nm and 530 nm light were used for imaging with LED driving currents of 150 and 80 mA, respectively). Citrine and EYFP expression were imaged with a GFP filter cube (Semrock, FF02-470/30 excitation, FF01-520/35 emission and FF495-Di03 dichroic) and tdTomato was imaged with a TRITC filter cube (Semrock, FF01-543/22 excitation, FF01-593/40 emission and FF562-Di03 dichroic).

For most channelrhodopsin stimulation experiments 10 mW/ mm2 of 470 nm and 590 nm stimulation lights from Thorlabs 4 channel LED light source (LED4D118) were delivered to the specimen plane. The light intensity was controlled with a Thorlabs DC4104 4 channel controller electronically by the separate analogue outputs of a HEKA ITC-18 DAQ board under the control of WinWCP 5.5 software. In the spectral response experiment, a Sutter VF-5 system with a smart shutter in the neutral density mode was used to achieve cross spectrum stimulation at the same photon flux at 1.586 x 1015 photons / s / mm2 (25.1 mW/mm2 at 400 nm and 15.4 mW/mm2 at 650 nm) between 400-650 nm (25 nm increments). The VF-5 has the following filters and bandwidth settings: 451-398 nm with 15 nm bandwidth, 503-446 nm with 15 nm bandwidth, 564-498 nm with 14 nm band-width, 632-557 nm with 14 nm bandwidth, 703-627 nm with 13 nm bandwidth. A second Sutter shutter was placed in the light path before the light guide to control the light output. For experiments with ACRs, a 15-sec interval was used between stimulation episodes. For ChrimsonR/IvfChrimson stimulation, stimulation light pulses either have 15-sec intervals between stimulation episodes (no preconditioning) or a 15-sec interval -1-sec 405 nm light pulse – 15-sec interval stimulation pattern for preconditioning before the subsequent stimulation. The stimulation light was guided into the fluorescent excitation light path of the Olympus IX73 microscope as with imaging but reflected with a Semrock FF685-DiO2 dichroic mirror to the objective.

Light intensity (in mW/mm2) was measured with a Thorlabs slide based light detector (S170C) at the specimen plane with illumination area defined by the microscope aperture and manually calculated for photon flux or illumination intensity.

For comparison of the membrane expression of our Ivf-Chrimson-citrine and ‘2A’, transfected cells were imaged on an Olympus BX51 microscope with a 40x Olympus water immersion objective, X-Cite 110LED and Hamamatsu Orca Flash4. The images were taken at 15% light intensity and 250ms exposure time with FF02-472/30 excitation filter, FF01-520/35 emission filter and FF495-Di03 dichroic mirror (Semrock, NY). A line profile was drawn across the cell and the membrane values were averaged and used for calculating the ratio with the mean cytosolic values.

Organotypic slices and virus infection

All procedures involving animals were approved by the Danish Animal Experiment Inspectorate. Hippocampi were isolated from postnatal P5-6 wild-type Sprague-Dawley rats (Janvier). The pups were decapitated and the brain removed and kept in an icy low sodium ACSF solution composed (in mM) of 1 CaCl2, 4 KCl, 1 MgCl2, 26 NaHCO3, 10 D-Glucose, 234 Sucrose, and 0,1% Phenol Red Solution, in mQ H2O, and bubbled with a mixture of CO2 (5%) and O2 (95%). From that moment, all the steps were performed in a laminar flow tissue culture hood using sterile equipment and in total asepsie. The brain was poured into a petri dish filled with the low sodium ACSF solution for hippocampi dissection under microscope guidance (Stereomicroscope, Olympus x1 objective). After extraction, the hippocampi were sliced at 400um thickness using an automatic tissue chopper (McIL-WAIN) and moved to a dish with pre-heated culture medium containing MEM Eagle medium 78.8% (Gibco), 20% heat-inactivated horse serum (Gibco), 1 mM L-glutamine, 1 mM CaCl2, 2mM MgSO4, 170 nM insulin, 0.0012% ascorbic acid, 12.9 mM D-glucose, 5.2 mM NaHCO3, 300 mM HEPES (Sigma), pH = 7.2-7.3, osmolarity adjusted to 315–325. The slices with intact DG and CA regions were then transferred onto air-fluid interface-style Milli-cell culture inserts (Millipore) in 6-well culture plates (ThermoFisherScientific) with 800 uL of sterile medium added below each insert. The slices were kept in a sterile incubator at 37 degrees with 5% CO2 (ThermoScientific, Steri-cycle i-160). The medium was replaced by a pre-warmed medium (37 degrees) three times a week.

After two to three days of culture, the slices were microinjected in DG (or in CA1 when specified) along the characteristic horseshoe pattern with a pulled glass pipette containing the following viruses: retro-vfChrimson (titer: 7×1012 particles/mL), retro-Zip(151T) (titer: 9×1012 particles/mL), retro-ZipT-IvfChr (titer: 2×1012 particles/mL), or retro-ZipV-IvfChr (titer: 1.5×1012 particles/mL). The injections were done under a microscope (using a Picospitzer III (Parker) connected to a Pulse Pal (10 ms pulse, every 0.5-sec).

Whole-cell electrophysiology and light delivery in-vitro

Two to three weeks post-infection, the organotypic slices were transferred into the recording chamber and continuously perfused with Artificial Cerebrospinal Fluid containing (in mM): 119 NaCL, 2.5 KCl, 26 NaHCO3, 1 NaH2PO4, supplemented with 11 D-glucose, 4 CaCl2, 4 MgCl2. Additionally, APV (50mM), NBQX (100mM) and picrotoxin (100mM) were added to block excitatory and inhibitory fast transmission. The solution was adjusted to pH 7.4 (osm ∼330) and bubbled with a mixture of CO2 (5%) and O2 (95%), at room temperature. The chamber was mounted on an upright microscope (Scientifica) linked to a digital camera (QImaging Exi Aqua) and the cells were visualized using 60X water-immersion objective (Olympus, LumiPlan). Acquisitions were performed in whole-cell configuration using Clampex 10.6 connected to a Multiclamp 700B amplifier via a Digidata 1550A digitizer (all from Molecular Devices). Voltage-clamp data were low-pass filtered at 200 Hz and digitized at 10 kHz and the whole-cell capacitance was compensated. Patch pipettes (2-4 MΩ of resistance) were filled with an internal solution containing (in mM): 153 K-gluconate, 10 HEPES, 4.5 NaCl, 9 KCl, 0.6 EGTA, 2 MgATP, and 0.3 NaGTP. The pH and osmolarity of the internal solution were close to physiological conditions (pH 7.4, osmolarity 297 mOsm). The access resistance of the cells in our sample was ∼ 25 MOhm. For voltage-clamp experiment, the Dentate Gyrus granule cells (otherwise specified when done in CA1 pyramidal cells) were clamped at -60mV. For current-clamp experiments, the cells were maintained at -60mV (except for the protocol where the potential was brought from -70 to -40mV) by constant current injection (holding current ± 100pA).

The recorded neurons were illuminated with 470 nm and 635 nm (nominal wavelengths) light with a coolLED pE-4000 system connected to the Digidata via a single TTL. The light irradiance was controlled via the control pod and was adjusted to 4mW/ mm2 for 470 nm and 7 mW/mm2 for 635 nm (otherwise specified). Light irradiance was measured with a Thorlabs digital optical power meter (PM100D) at the specimen plane with illumination area defined by the microscope aperture. Between the recordings, wavelengths were switched from 470 nm to 635 nm and vice versa. At the end of the recordings, the location of the recorded cell was confirmed by inspection at 10x magnification. Data analysis was performed using Clampfit 10.6 (Molecular device).

After recordings, slices were kept in 4% formalin, and later mounted under coverslips with Fluoromount (Sigma-Aldrich). Pictures were taken with an Apotome.2 (Zeiss) and ZEN software at 10X and 40X magnifications.

Acute slices

Experimental procedures were approved by the Animal Care and Use Committee of the University of Buenos Aires (CICUAL). Briefly, 4-week old C57 mice (n = 4) were co-injected with AAV-8/2-hSyn1-ChrimsonVF_EYFP_Kv2.1 and AAV-8/2-hSyn1-ZipACR_Kv2.1_IRES_DsRed2 within the dentate gyrus of the hippocampus (250nl of a 1:1 mixture, coordinates from Bregma: 1 mm mediolateral, -2 mm anteroposterior, -1.5 dorsoventral). After 3 weeks of expression, the animals were sacrificed and the brain removed and cut in a solution composed of (in mM): 234 sucrose, 11 glucose, 26 NaHCO3, 2.5 KCl, 1.25 NaH2PO4, 10 MgSO4, and 0.5 CaCl2 (equilibrated with 95% O2–5% CO2). The slices were maintained at room temperature before being transferred in a recording chamber mounted on a microscope (Nikon) connected to a Mightex Illumination system for 470 nm and 635 nm light delivery (at 10mW). For current clamp experiments the amount of current injected was corrected in the inter-sweep interval to keep it close to -60 mV. In voltage clamp experiments holding potential was -60 mV.

Recordings were done in a solution of composition (in mM): 126 NaCl, 26 NaHCO3, 2.5 KCl, 1.25 NaH2PO4, 2 MgSO4, 2 CaCl2 and 10 Glucose (pH 7.4).

Patch-pipettes (2-4 MΩ of resistance) were filled with a K-gluconate based internal solution (in mM, 130 K-gluconate, 5 KCl, 10 HEPES, 0.6 EGTA, 2.5 MgCl2·6H2O, 10 Phosphocreatine, 4 ATP-Mg, 0.4 GTP-Na3).

Data Analysis

The data were analysed and plotted using GraphPad Prism V9 (GraphPad Software, La Jolla, CA, USA). All values are indicated as mean ± standard deviation (SD) or mean ± standard error of the mean (SEM), as specified . For statistical comparison between groups, data were tested for normal distribution using D’Agostino and Pearson, Shapiro-Wilk or KS normality test. Comparison of the mean between groups has been done with a one-way analysis of variance (ANOVA) followed by Dunnett’s multiple comparisons test.

ZEN software (Zeiss) and FIJI (from ImageJ) were used for images processing.

Figure 1A, 1B1 and 3E1 have been created with BioRender.com.

eLife Assessment

This study develops valuable tools for optogenetic control of neuronal activity. The basic characterization of the activation of a red-shifted channelrhodopsin paired with a blue-light sensitive anion channel engineered to obtain desired inhibitory current kinetics is solid. However, guidelines and feasibility for their practical use under simultaneous multi-color stimulation are incomplete, as are comparisons with current state-of-the-art tools.

Reviewer #1 (Public Review):

Summary

In this manuscript, the authors generate an AAV-deliverable tool that generates action potentials in response to red light, but not blue light, when expressed in neurons. To do this, they screen some red light-excitatory/blue light-inhibitory opsin pairs to find ones that are spectrally and temporally matched. They first show that this works with Chrimson and GtACR2, however, they expand their search after finding that the tau-off (inactivation after light cessation) kinetics of these two opsins are not well-matched. They directly examine a small set of options based on a literature search and settle on a variant of red light-excitatory Chrimson and blue light-inhibitory ZipACR. To more closely match the kinetics of this pair, the authors create a structure homology model of the ZipACR retinal binding pocket and use this to guide the generation of a small mutagenesis panel, leading to a more optimized ZipACR mutant. They then show that a bicistronically expressed fusion arrangement of these opsins, plus some functional peptides, can drive action potentials up to 20hz with red light and does not do so with blue light, in hippocampal cells transduced by AAV. They also show function in vivo, in a mouse, using a physiological readout. They conclude that their new tool may be useful for complex experimental designs requiring multiple optical channels for write-in/read-out.

The major advantage claimed by the authors over existing tools is the temporal time-locking of their inhibitory opsin - this is driven by the contrast between the tau-off kinetics of their ZipACR variant compared to gtACR2, which is used by the leading competitor tool (BiPOLES).

Big thoughts

While the authors were carefully thoughtful about the potential influence of temporal kinetics on the efficiency of a tool such as this one, there were no experiments conducted that made use of the unique properties of this molecular strategy. To understand why they embarked on this engineering program, I was required to put on my neuroscientist hat and contemplate this question myself:

First, experimental designs where I require multiple optical channels of control. This appears to be aligned with the author's thoughts, as they state, correctly, that opsins utilizing retinal as a light-sensing chromophore are universally activated by blue light (the so-called 'blue shoulder'). Therefore, their tool may be useful for stimulating multiple populations using a blue excitatory opsin in neuron A and their tool for red excitation of neuron B - or, in the author's own words, "A potential solution to the problem of cross-talk...". Yet, there are no data presented that showcases their new tool for this purpose (e.g. Vierock, Johannes, et al. "BiPOLES is an optogenetic tool developed for bidirectional dual-color control of neurons." Nature Communications 12.1 (2021): 4527. Figure 4f-I; 6). The same set-up could be imagined for green GECI (or equivalent) imaging of cells in the same volume that their tool is being used in - for instance, interleaving red stimulation light and blue imaging light, (perhaps) without the typical concern of imaging light bleed-through activating the opsin itself.

Second, for high-frequency temporal control over both excitation and inhibition in the same neuron. The red light turns the cell on, and blue light turns the cell off (see, for instance, Zhang, Feng, et al. "Multimodal fast optical interrogation of neural circuitry." Nature 446.7136 (2007): 633-639. Figure 2; Vierock as above, Figure 4a,b). Again, here the authors are long on theory ("The new system...can drive time-locked high-frequency action potentials in response to red pulses") and short on data. While they do show that red light = excitation and blue light = inhibition, they neither show 1) all-optical on/off modulation of the same cell; nor 2) high-frequency inhibition or excitation (max stim rate of 20hz, which is the same as the BiPOLES paper used for their LC stimulation paradigm; Vierock, as above, Figure 7a-d).

Despite these major shortcomings, the further development and characterization of tandem opsins, such as this one, is of interest to the community. There is ongoing work by the BiPOLES team to create new iterations (e.g. Wahid, J., et al. "P-15 BiPOLES2 is a bidirectional optogenetic tool with a narrow activation spectrum and low red-light excitability." Clinical Neurophysiology 148 (2023): e16.). To make the case that the tool described in this manuscript is worth the effort that the authors are requesting the neuroscience community invest in trialing it in their own hands, they must revise the manuscript to show that their approach is both 1) different in some way when compared to BiPOLES (it is my understanding that they did not do this, as per the supplementary alignment of the BiPOLES sequence and the sequence of the BiPOLES-like construct that they did test) and 2) that the properties that the investigators specifically tailored their construct to have confer some sort of experimental advantage when compared to the existing standard.

There are a number of additional concerns and clarifications that will strengthen the manuscript that are communicated directly to the authors through this peer-review process.

Reviewer #2 (Public Review):

Summary:

One often wishes to combine activation of a neural population via red light with simultaneous modulation of a different population via blue light, or simultaneous imaging of a blue-excited fluorescent reporter. The problem is that all red-shifted opsins have an action spectrum with a long blue tail, leading to spurious opsin activation by blue light.

This valuable paper presents a clever solution to this problem, by pairing an engineered blue-shifted inhibitory chloride-conducting opsin with a red-shifted excitatory opsin. The combined effect is excitation by red light and shunting inhibition by blue light. The paper is very thorough, with convincing spectroscopic and patch clamp characterization of the tools, and tests in brain slices and in vivo. This tool is likely to be useful in the neuroscience community.

Strengths:

The methods are solid, including the complete characterization of each tool separately, as well as the combination in vivo. The array of testing gives a strong degree of confidence that this tool will work as expected.

Weaknesses:

There are two discussion points and one experimental point which would make the paper stronger.

1. In the Introduction or Discussion, the authors could better motivate the need for a red-shifted actuator that lacks blue crosstalk, by giving some specific examples of how the tool could be productively used, e.g. pairing with another blue-shifted excitatory opsin in a different population, or pairing with a GFP-based fluorescent indicator, e.g. GCaMP. The motivation for the current tool is not obvious to non-experts.

2. Simultaneous excitation and inhibition are not the same as non-excitation. The authors mentioned shunting briefly. Another possible issue is changes in osmotic balance. Activation of a Na+ channel and a Cl- channel will lead to net import of NaCl into the cell, possibly changing osmotic pressure. Please discuss.

3. The authors showed that in ZipT-IvfChr, orange light drives excitation and blue light does not. But what about simultaneous blue and orange light? Can the blue light overwhelm the effect of the orange light? Since the stated goal is to open the blue part of the spectrum for other applications, one is now worried about "negative" crosstalk. Please discuss and, ideally, characterize this phenomenon.

Reviewer #3 (Public Review):

This study addresses the important topic of dual-color optogenetic control of neuronal activity, which is challenging due to significant optical crosstalk between channelrhodopsins of different absorption colors and ion selectivity. However, Mermet-Joret et al. demonstrate in flies that simple coexpression of a strong blue light-activated inhibitory opsin, such as the chloride-selective channelrhodopsin GtACR2, can suppress the blue light activity of a red-shifted excitatory opsin, such as Chrimson, and allow dual-color optogenetic control of the expressing neuron. The same concept was previously discussed by Vierock et al. and led to the generation of BiPOLES, which combines both channels in a single fusion protein. In the present manuscript, the authors introduce an alternative combination of channels with accelerated off-kinetics that are coexpressed by a bicistronic expression cassette. The goal is to better match the duration of illumination and optogenetic manipulation in order to reduce potential side effects induced by prolonged channel opening.

The major novelty of this work lies in the choice of the employed ion channels: the excitatory cation channel vf-Chrimson and the inhibitory anion channel ZipACR, alongside their subsequent modifications (Fig. 2 - 4). Both channels belong to the fastest known ChRs, but the choice of ZipACR raises questions. First, it has a peak absorption at 515 nm that is 40 nm further red-shifted than GtACR2 tested in Figure 1 and accordingly important optical cross-talk with the coexpressed Chrimson channel. Second, it was reported to have reduced chloride selectivity, first by Govorunova et al. in 2017 and later also by Kato et al. in 2018. Both of these aspects are also mentioned by the authors but were not resolved through molecular engineering. Instead site-directed mutagenesis primarily focused on membrane expression and photoreceptor kinetics of the employed channels. Nonetheless, improving the membrane targeting of the vf-Chrimson channel by exchange of the N-terminus finally provided sufficient red light activation at low light intensities to reliably activate expressing neurons and allowed in combination with the decelerated ZipACR mutants dual color optogenetic control with millisecond time resolution. At higher light intensities inactivation of Chrimson and the optical crosstalk of both channels seem to limit its performance.

The experimental results are well presented; but, certain questions persist:

1. The enhanced vf-Chrimson could potentially be a highlight of the manuscript, serving broader applications. Yet, gauging the overall improvements of ivf-Chrimson in comparison to other Chrimson variants remains intricate due to several reasons. First, photocurrents from ivf-Chrimson seem smaller than those from C-Chrimson (Supplemental Figure 3), and a direct comparison with standard vf-Chrimson is absent. Second, while membrane expression of ivf-Chrimson appears enhanced in provided bright-field recordings, the quantitative analysis would necessitate confocal microscopy and a membrane marker (Supplemental Figure 2). Finally, other N-terminal modified Chrimson variants, like CsChrimson by Klapoetke et al. in 2014 and C1Chrimson by Oda et al. in 2018, have been generated. Comparing ivf-Chrimson to vf-CsChrimson or vf-C1Chrimson would be important to evaluate the benefits of the applied N-terminal modification.

2. The action spectra of ZipACR suggest peak absorption of ZipACR WT and its mutant at 525 - 550 nm (Fig. 3). This is even further red-shifted than previously reported by Govorunova et al. Further action spectra recordings differ for all constructs between recordings initiated with blue or red light (Supplementary Fig. 5). This discrepancy is unexpected and should be discussed. Additionally, the representative photocurrents of Zip(151V) in Fig. 3D1 do not align with the corresponding action spectrum in Fig. 3D2 as they show maximal photocurrents for 400 nm excitation.

3. The authors introduce two different bicistronic expression cassettes-ZipT-IvfChR and ZipV-IvfChR-without providing clear guidelines on their conditions of use. Although the authors assert that ZipT is slower and further red-shifted than ZipV, the differences in the data for both ACR mutants are small and the benefits of the different final constructs should be explained.

4. The ZipT/V-IvfChRs are designed as bicistronic constructs; yet, disparities in membrane trafficking and protein degradation between the two channels could lead to divergences in blue and red light photoresponses. For future applicants, understanding the extent of expression ratio variations across cells using the presented expression cassettes could be of significance and should be discussed.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Therefore, their tool may be useful for stimulating multiple populations using a blue excitatory opsin in neuron A and their tool for red excitation of neuron B… Yet, there are no data presented that showcases their new tool for this purpose

We agree with the reviewer that in this manuscript we have not experimentally shown the applicability of our system for dual optical stimulation. However, the suppression of blue-light excitation of ZipV/T-IvfChr-expressing neurons strongly suggests this can be used in experiments exciting populations of neurons similarly shown for BiPOLES. We don’t see a theoretical basis where this experiment cannot be done if sufficient cell targeting mechanisms (such as the use of cre-lox or retroAAV) is utilized. We have started several projects pursuing these utilities in the meantime.

While they do show that red light = excitation and blue light = inhibition, they neither show 1) all-optical on/off modulation of the same cell; nor 2) high-frequency inhibition or excitation (max stim rate of 20hz, which is the same as the BiPOLES paper used for their LC stimulation paradigm; Vierock, as above, Figure 7a-d).

Regarding point 1, we understand that the reviewer asks if we have optically excited (with red light) and inhibited (with blue light) the same neurons. If so, figure 4B1 (optical excitation of ZipT-IvfCh with red light) and figure 5A (optical inhibition of ZipT-IvfCh with blue light) represent largely the same set of neurons.

Regarding point 2, we respectfully disagree with the reviewer’s interpretation of Figure 7a-d) in Vierock et al. As we understand, in this part the authors apply a 20 Hz optical stimulation protocol to the LC neurons in vivo. However, there is no data showing that individual neurons do follow this stimulation protocol. To be clear, we are not saying that BiPOLES cannot drive 20 Hz APs. Very likely it can. It is based on ChrimsonR which is capable of doing so (Klapoetke et al., Figure 2). Although, in this manuscript we have not shown data for optical stimulation above 20Hz, our system is based on vfChrimson, which is known to drive AP of 100Hz and above (Mager et al., figure 2 and 3).

… they must revise the manuscript to show that their approach is both 1) different in some way when compared to BiPOLES (it is my understanding that they did not do this, as per the supplementary alignment of the BiPOLES sequence and the sequence of the BiPOLES-like construct that they did test) and 2) that the properties that the investigators specifically tailored their construct to have confer some sort of experimental advantage when compared to the existing standard.

In the latest version of the manuscript, we have compared our ZipV-IvfChr and the BiPOLES construct adapted with vfChrimson (Fig. 2 Suppl 1). The mean photocurrent amplitude of IvfChr in the ZipV-IvfChr construct is ~2.7 x higher than BiPOLES adapted with vfChrimson (14 randomly selected HEK293 cells in each group) (Fig. 2 Suppl 1B). We conducted this experiment in HEK293 cells to ensure accurate voltage-clamping and less biased cell selection. Even adjusting for the smaller photocurrent of vfChrimson vs ChrimsonR, this would still translate to ~1.6 x greater photocurrent with ZipV-IvfChr compared to the original BiPOLES utilizing ChrimsonR. We believe the increased efficiency of excitation is an important aspect of adapting vfChrimson for red-light excitation of neurons.

Reviewer #2 (Public Review):

(1) In the Introduction or Discussion, the authors could better motivate the need for a red-shifted actuator that lacks blue crosstalk, by giving some specific examples of how the tool could be productively used, e.g. pairing with another blue-shifted excitatory opsin in a different population, or pairing with a GFP-based fluorescent indicator, e.g. GCaMP. The motivation for the current tool is not obvious to non-experts.

In the discussion, we now provided examples for potential use of the tool. For example, one of the key aspects that can be manipulated by the existing tool is the induction of spike-timing dependent plasticity with 2 wavelengths of light with blue light channelrhodopsin such as oChIEF is used to evoke presynaptic release and ZipT-IvfChr expressed in postsynaptic neuron. In this situation, the rapid termination of inhibitory response is critical so it does not interfere with the induction of LTP or LTD. Another experiment is the alternate control of projection neurons and interneurons in cortical areas, independent controls of neurons of direct and indirect pathways in the striatum to manipulate behavior.

(2) Simultaneous excitation and inhibition are not the same as non-excitation. The authors mentioned shunting briefly. Another possible issue is changes in osmotic balance. Activation of a Na+ channel and a Cl- channel will lead to net import of NaCl into the cell, possibly changing osmotic pressure. Please discuss.

We agree with the notion that osmotic, ionic and pH changes in small neuronal structure can be disruptive to the physiology and this is the reason we developed our approach where the fastest channelrhodopsins are used so we can minimize the channel opening time and the flux of ions through the channels when brief light illuminations are applied. Not only the flux of protons, sodium ions and calcium ions are minimized, the flux of chloride should be minimal as well (as the membrane potential should be close to the reversal potential of chloride reversal potential hence low ion flow). Hence our approach should be minimally disruptive compared to most other existing channelrhodopsin-based approaches when short or minimal light pulses were used in conjunction with our tools. This recommendation is included in the updated manuscript .

(3) The authors showed that in ZipT-IvfChr, orange light drives excitation and blue light does not. But what about simultaneous blue and orange light? Can the blue light overwhelm the effect of the orange light? Since the stated goal is to open the blue part of the spectrum for other applications, one is now worried about "negative" crosstalk. Please discuss and, ideally, characterize this phenomenon.

We now have performed this experiment. Simultaneous blue (470nm) and red light (635nm) stimulation does not produce AP (Fig .4 Suppl 1A)). This suggests the inhibitory effect of ACR is more efficient than the excitatory effects of IvfChr due to their higher conductance, this re-emphasizes the rapid termination of the ACR effects is critical for minimal disruption of physiological effects in such pairing strategy.

(3.1) Does the use of the new tool require careful balancing of the expression levels of the ZipT and the IvfChr? Does it require careful balancing of blue and orange light intensities?

As with any optogenetic tool, the users should validate the efficacy of the tool in their own system. Our tool solely relies on the balanced expression of the 2A system, the efficiency of the two opsins and their degradation of the time-span of expression. These aspects of the tool would be better addressed in future versions of the tools or improvement of the BiPOLES-type of tandem expression in subsequent versions. From the instrumentation side, the light intensity and differential penetration depth requires careful consideration. However, this holds true in most optogenetic and fluorescence imaging-based approaches as well. In the current update of the manuscript, we have included further discussion on these aspects as well.

(3.2) Also, many opsins show complex and nonlinear responses to dual-wavelength illumination, so each component should be characterized individually under simultaneous blue + orange light.

We now have performed this experiment (please see our comment to point 3)

(3.3) I was expecting to see photocurrents at different holding potentials as a function of illumination wavelength for the coexpressed construct (i.e. to see at what wavelength it switches from being excitatory to inhibitory); and also to see I-V curves of the photocurrent at blue and orange wavelengths for the co-expressed constructs (i.e. to see the reversal potential under blue excitation). Overall, the patch clamp and spectroscopic characterization of the individual constructs was stronger than that of the combined constructs.

We have added the IV curves for the co-expressed construct at different holding potentials for 470nm and 635nm wavelengths. This shows reverse potential for the two wavelengths that are intended for in vitro and in vivo applications. Performing a similar experiment for a variety of wavelengths would not be as valuable, in part, due to the enormous amount of data generated. As we have shown in the study, the response of any channelrhodopsins vary with different light duration and light intensities in addition to the wavelengths and holding potentials. The results for each recorded cell could include stimulation by different wavelengths, stimulation by different illumination intensities, stimulation with different light duration in addition to different holding potentials. Not only would the results be highly variable from cell-to-cell, there will be potentially hundreds or thousands of combinations to be tested per cell (e.g., 5 light intensities @1, 2.5 , 5 , 10 and 20 mW/mm>sup>2, 8 different wavelengths @ 450nm, 475nm, 500nm, 525nm, 550nm, 575nm, 600nm and 625nm, 7 light durations @ 1ms, 5ms, 10ms, 50ms, 100ms, 500ms and 1s, and , and 6 holding potentials @ -80mV, -70mV, -60mV, -40mV, -20mV and 0mV would result in 1680 stimulation conditions per recorded cell).Technically, the significant lowering of membrane resistance when both IvfChr and ZipACR variants are activated simultaneously would compromise the quality of voltage-clamping even in HEK293 cells with series resistance compensation. We have yet to see any other studies that had included such ambitious electrophysiology experiment for the channelrhodopsin characterization, likely due to the feasibility of such experiment.

Reviewer #3 (Public Review):

(1) The enhanced vf-Chrimson could potentially be a highlight of the manuscript, serving broader applications. Yet, gauging the overall improvements of ivf-Chrimson in comparison to other Chrimson variants remains intricate due to several reasons. First, photocurrents from ivf-Chrimson seem smaller than those from C-Chrimson (Supplemental Figure 3), and a direct comparison with standard vf-Chrimson is absent.

We appreciate the reviewer’s positive view of our modified variant. We did not emphasize this particular modification as it was identical to our previous published modification and similar to that previously published by others (CsChrimson and C1Chrimson). In all these cases, improved membrane expression was consistently detected. We believe that expression data and our comparison of C-Chrimson and IvfChr is sufficient to justify the improved membrane expression and function.

Second, while membrane expression of ivf-Chrimson appears enhanced in provided brightfield recordings, the quantitative analysis would necessitate confocal microscopy and a membrane marker (Supplemental Figure)

We have now quantified the results with a membrane palmitoylated mCherry using confocal microscopy shown in Fig 2 Suppl1 A. We measured the Pearson Correlation Coefficient of the mCherry with EGFP or Citrine signal for the 6 constructs (vfChrimson, vfChrimson with trafficking sequence, vfChrimson with N-terminal signaling peptide from oChIEF (C-vfChrimson), vfChrimson with trafficking sequence and N-terminal signaling peptide from oChIEF (IvfChr), BiPOLES with EGFP or citrine and vfChrimson) and the results were identical and consistent with the prior results using epifluorescence microscopy.

(2) Finally, other N-terminal modified Chrimson variants, like CsChrimson by Klapoetke et al. in 2014 and C1Chrimson by Oda et al. in 2018, have been generated. Comparing ivf-Chrimson to vf-CsChrimson or vf-C1Chrimson would be important to evaluate the benefits of the applied N-terminal modification.

Our development of IvfChrimson is similar to the approach of vf-CsChrimson and identical to that of vf-C1Chrimson and we do not claim these modifications to be unique or superior. However, we have developed our design independently of these other studies and we have more extensive functional comparison and characterization data of our IvfChrimson variant than the other studies.

(2.1) The action spectra of ZipACR suggest peak absorption of ZipACR WT and its mutant at 525 - 550 nm (Fig. 3). This is even further red-shifted than previously reported by Govorunova et al. Further action spectra recordings differ for all constructs between recordings initiated with blue or red light (Supplementary Fig. 5). This discrepancy is unexpected and should be discussed.

We thank the reviewer for the comment, this was a mistake in the traces used for the figure. The example traces were the spectral response measured from the 400 nm to 650 nm instead of the 650 nm to 400 nm order shown in the spectral data. This has now been corrected.

Additionally, the representative photocurrents of Zip(151V) in Fig. 3D1 do not align with the corresponding action spectrum in Fig. 3D2 as they show maximal photocurrents for 400 nm excitation.

Please, see point above.

(3) The authors introduce two different bicistronic expression cassettes-ZipT-IvfChR and ZipV-IvfChR-without providing clear guidelines on their conditions of use. Although the authors assert that ZipT is slower and further red-shifted than ZipV, the differences in the data for both ACR mutants are small and the benefits of the different final constructs should be explained.

In our testing in neurons, ZipT has less ‘escaped’ spikes after the termination of the light pulses in the cells we have tested. However, this is dependent on the membrane properties such as capacitance and resistance of the cells. ZipV has a faster termination time and in some situations may be necessary due to its faster termination time and reduced disruption of physiological processes.

We have now included this discussion in our updated manuscript.

(4) The ZipT/V-IvfChRs are designed as bicistronic constructs; yet, disparities in membrane trafficking and protein degradation between the two channels could lead to divergences in blue and red light photoresponses. For future applicants, understanding the extent of expression ratio variations across cells using the presented expression cassettes could be of significance and should be discussed.

We now have included this discussion in our responses above.

Reviewer #1 (Recommendations For The Authors):

(1) The Figure 1a mV cartoon traces for chloride are confusing. The chloride currents are depolarizing, not hyperpolarizing. As noted by the authors, these channels largely generate AP blockade through shunting inhibition (division), not hyperpolarization (subtraction).

The figure has been corrected.

(2) Figure 2A does not show where the light is applied. Why are some of the bars blue and some of them not filled?

This has been corrected

(3) Figure 2C1 does not show where the light is applied. There should be an inset to detail the blue-light-cessation-evoked AP. Also doesn't give the holding potential.

The requested details are added.

(4) Figure 2C2 inset is described as showing that "Light-induced currents with 470 nm illumination were initially outward but turned inward immediately following light offset." Is that correct? It looks to me like the current turns inward about half-way through the light pulse and then becomes even stronger after the light turns off. That is also consistent with the CC traces, which appear to show a transition toward depolarization during the light pulse before the AP initiation at light offset.

Yes, the reviewer's observation is correct. There are blue light-induced outward and inward current peaks at the onset and offset of the light. Accordingly, we have modified the phrasing for Fig. 2C2.

(5) Figure 3D1 shows that Zip(151V) has a peak current at 400nm, with a steady increase in current from red to blue, however, this is not the case in the summary data in 3D2. It's also not shown in Supplementary Figure 5B. What's going on?

We apologize for the prior version of the figure associated with the first submission. The example traces from 400nm -> 650 nm were incorrectly included in the figure whereas the 650nm -> 400 nm example traces should be included. This has been corrected.

(6) Figure 3D1 has no time scale.

It is now been included

(7) Figure 3E1 should read "Transduced" and not "Transfected"

This has been corrected.

(8) IvfChr fidelity drops off dramatically at 20hz...down to 50% efficiency of generating APs. This is described in the legend as "high frequency". Maybe the cart came before the horse in this figure...as it looks like in panel C that using less light power density improves fidelity in the dual opsin configuration with red light stimulation...why not use that power for the characterization? Did you try any higher frequencies? Or longer pulse widths? This is an important characterization to inform further use of the tool. This shortcoming isn't a cell-intrinsic limitation, as the 470nm stim with IVfChr was 100% successful at both 10hz and 20hz.

It is known that red but not blue light pulses induce desensitization (optical fatigue) in red-shifted ChR variants. Indeed, one can reinstate the response to red light, by giving violet-blue light pulses (Fig 4. Suppl 2). We think this is the reason that the 470nm stimulation was more effective in inducing AP in cells expressing IvfChR. Higher light intensities induce greater desensitization, but are preferred for faster opening of channels and depolarization of neurons. This can explain why, in some situations, lower light intensities were more effective in producing APs when pulse trains were used. We have recordings from cells firing APs at 40Hz (not included). All these cells had high expression levels of the opsin.

(9) Figure 4D: why use 100ms pulse width? How do you know that this isn't causing depol block? Or some of the nefarious concerns that are raised in the discussion, such as "...disrupt[ion of] normal neuronal physiology and signal processing that occurs in millisecond time scale"?

We used 100ms pulse duration to follow the published protocol that this experiment is based on (Lin et al., 2013, Nature Neuroscience).

(10) Figure 4E-bottom: What is the blue peak at light onset? Is the tool driving early activation before silencing?

There seems to be an early, sharp and brief activation by blue light. We don’t know the definite cause of this, but we speculate this is driven by blue-light activation of ZipACR and not the IvfChr portion of the construct. The reason is that such a sharp rise is absent when only IvfChr is expressed (Fig. 4E, upper panel). Soma-targeted motif tethered to channelrhodopsins is known to result in preferential expression of channels close to soma but does not exclude the expression of channelrhodopsin in axonal and dendritic compartments, especially when animals are allow to recover for long period of time after viral injection. We believe that ZipACR at axonal terminals where the chloride concentration is high can still cause blue-light evoked depolarization and transmitter release. We observed this phenomenon in two mice in their first trial. The data for individual trials for each mouse are included in a supplementary table.

(11) Figure 4G: Earlier in this same figure (B2, C), 470nm light was more effective at stimulating IvfChr than 635nm light. Is it unexpected that 638nm light would in this in vivo context be more effective at driving IvfChr responses than 450 nm light (at least as reflected by the AUC measurements)? Does this reflect fiber placement and light penetration/scattering?

The spectral peaks of Chrimson-based variants including vfChrimson are all centered around 600 nm, and at 635 / 638 nm light, the amplitudes of photo-response decline, the channel onset slows, and the channels suffer greater desensitization. In isolated preparations where the light penetration is similar between 635 / 638 nm and 470 nm, 470 nm responses can outperform 635 / 638 nm responses due to its lack of desensitization and higher consistency in its response. This is also a strong reason that we have developed our current approach. In in vivo preparation shown in Fig. 4D-G, the much higher tissue penetration of 638nm light due to reduced absorption and reduced scattering can offset the performance of IvfChr to 450 nm light.

(12) In the methods, it is noted that different viral batches appear to generate different levels of neuronal toxicity. If that is the case, how did you differentiate between true differences between constructs vs. differential cell health effects?

For figure 4D-F (whisker movement), we determined virus toxicity using NeuN staining. In slice recordings, we used the electrophysiological property of the neurons to assess their health. For this manuscript, we had one batch of virus that produced toxicity. We did not include any data from this batch.

Reviewer #2 (Recommendations For The Authors):

● Define AUC on first use.

It is now defined.

● Figure 3C2: Please explain how the photocurrents were normalized. As presented, it looks like under strong orange light, the ZipACR has higher photocurrent than the ivfChr.

This is due to the fact vfChrimson and other Chrimson-based variants do not fully recover in the dark after 590 nm stimulation. We tested IvfChrimson with both reconditioning light pulse of 405 nm and without 405 nm and we can consistently reach a greater ‘maximal’ response from the same cell after 405 nm reconditioning (see Fig. 4 Suppl 2). We therefore normalize the response to the maximal recorded response of the cell often achieved with 10 or 20 mW/mm2 590 nm stimulation after 405 nm reconditioning. We understand this can be confusing and have now replaced the light-intensity response in Fig. 3C2 with the one with 405 nm reconditioning which is easier to interpret for the readers.

● P. 3: "As expected, blue light pulses induce transient membrane suppression..." Unclear what "suppression" means. Shunting? Hyperpolarization?

We rephrased this to “As expected, blue light pulses transiently suppress APs…”

● P. 3: "illumination at 470 nm and 590 nm wavelengths led to similar amounts of courtship song (110.1 {plus minus} 12.8 and 78.5 {plus minus} 11.6,n = 16-17, respectively)". What are the units of "courtship song"?

The unit for courtship song is the number of pulses per 10 seconds. This has been clarified in the figure.

● P. 5: The quantification of photocurrent in terms of pA/pF/A.U. is non-standard. I understand the impetus to normalize by expression to give something proportional to per-molecule conductance, but a user cares about overall photocurrent. Please also give the real photocurrents, either pA or pA/pF.

We have provided the real photocurrent in pA or pA/pF where scientifically appropriate. To avoid selection and experimenter’s bias in our data, we did not set criteria for data elimination for cells with specific fluorescence intensity or photocurrent amplitude. Some resulting response can range from vary up to 20 folds from the same construct in many experiments. We do not believe that averaging absolute photocurrent amplitude would be justified due to the imbalance of weighing in the results. We do acknowledge that not selecting or eliminating data points would introduce higher noise in recordings with smaller responses but this is preferable over the selection or experimenter bias that is likely to be introduced otherwise.

● Please quote illumination intensities wherever possible.

● P. 7: why was the red light crosstalk into Zip(151T) tested at 635 nm instead of 590 nm? Isn't the relevant parameter 590 nm, since that will be used for the excitatory opsin?

In all our characterizations of the constructs using slice electrophysiology recordings, we used 635nm instead of 590nm. The reason is that compared to 590nm wavelength, at 635nm the photocurrent for Zip(151T) and Zip(151V) is significantly reduced (Fig. 3D1,D2).

● P. 10: "we examined the power at which responses to 470 nm and 635 nm lights induce APs in neurons expressing ZipT-IvfChr, ZipV-IvfChr, or IvfChr", but the preceding sentence says you didn't test the ZipT-IvfChr. This is confusing, please clarify.

The previous paragraph refers to the photocurrent recordings in HEK293 cells where our fast LED based illumination system is limited to 590 nm light, whereas the subsequent paragraph refers to the brain slice neuronal recordings. We have now emphasized the difference of the experiments in the rewrite.

● Fig. 4B1, top: Why don't the blue traces return to the same baseline after the stimulus epochs?

We observed this shift in baseline (~4mV more depolarized) in cells expressing IvfChR (or vfChR) only with blue light stimulation. This was observed in the neurons recorded in the CA1 as well (data not shown). There was no such a change following red light stimulation (Fig. 4B1). Therefore, this should not affect the applicability of our construct. The original paper introducing vfChR did not test the responses of their constructs to blue light. There could be another photocycle state that is activated stronger by 470nm than 590nm and it has a slow off-rate, but this is only a speculation from our side. It must be noted we did not observe such a phenomenon in cells expressing ChrimsonR (Fig. 1 Suppl 1C).

● Fig. S3B, right: The two colors are barely distinguishable on the graph. Consider more distinct colors and/or different symbols.

It has been changed accordingly.

● P. 15: "However, we do not recommend the use of orange light pulses, as we observed a significant photocurrent in this wavelength." Not clear what this is referring to. Which construct? Under which circumstances shouldn't one use orange light pulses? Where's the data showing this?

This is referring to Fig. 3D1,D2 and Figure 4 suppl Fig. 2 which show a normalized ~40-50% photocurrent at 590nm. Now in the text, the reference figures for the data are added.

Version 2

Abstract

A well-known phenomenon in the optogenetic toolbox is that all light-gated ion channels, including red-shifted channelrhodopsins (ChRs), are activated by blue light, whereas blue-shifted ChRs are minimally responsive to longer wave-lengths. Here, we took advantage of this feature to create a system which allows high-frequency activation of neurons with pulses of red light, while permitting the suppression of action potentials with millisecond precision by blue light. We achieved this by pairing an ultrafast red-shifted ChR with a blue light-sensitive anion channel of appropriately matching kinetics. This required screening several anion-selective ChRs, followed by a model-based mutagenesis strategy to optimize their kinetics and light spectra. Slice electrophysiology in the hippocampus as well as behavioral inspection of vibrissa movement demonstrate a minimal excitation from blue light. Of significant potential value, in contrast to existing tools, the system we introduce here allows high frequency optogenetic excitation of neurons with red light, while blue light suppression of action potentials is confined within the duration of the light pulse.

Introduction

Despite the short history of optogenetics, optical manipulation of neuronal activity has become an indispensable tool in the studies for functional analysis of brain circuits. The increasing needs for a more precise temporal and spatial manipulation of the neurons along with the untapped potentials of optogenetics have been driving forces in expanding the repertoire of the optogenetic toolbox. Main development among these expansions is the discovery of opsins of diverse excitation spectra, with the current excitatory opsins showing light sensitivity ranging from near UV to near far-red spectra (Yizhar et al. 2011; Deisseroth and Hegemann 2017; Marshel et al. 2019).

A major concern in the use of opsins with longer excitation spectra is that all opsins, in addition to their own excitation spectral peak, are activated by blue light, a problem known in the field as cross-talk or bleed-through. Attempts to create blue light-insensitive opsins will likely be unsuccessful due to the intrinsic blue absorption by the retinal molecule, the non-encoded chromophore, which captures light energy in channelrhodopsins (Waddell, Schaffer, and Becker 1973; Wang et al. 2012).

There have been several standouts attempts to overcome the problem of cross-talk. One uses a variant of blue-shifted ChR with fast kinetics in combination with a slow kinetics red-shifted ChR (R-ChR) (Klapoetke et al. 2014). The reasoning is that short pulses of blue light will activate only fast ChR. Since it requires a tight regulation of the protein expression as well as the light intensity, this method has not been adopted for in vivo application. Another approach is to expose axons expressing an R-ChR to prolonged photostimulation, during which axons become unresponsive to light pulses and cease to fire (Hooks et al. 2015; Faress et al. 2024). This method is limited to axonal regions as prolonged optical stimulation of the soma induces trains of action potentials. Additionally, prolonged stimulation of the axons of neurons with slow spike-frequency adaptation, such as fast-spiking interneurons, may result in repetitive activity rather than depolarization block (Hooks et al. 2015).

A potential solution to the problem of cross-talk is to co-express a blue-shifted anion channelrhodopsin (B-ACR) along with the R-ChR to ‘suppress’ the excitation of R-ChR by blue light. The intention is to have a neuron co-expressing such opsins selectively produce action potentials to red light and not blue light. A recent study achieved this, by pairing GtACR2 and Chrimson (Vierock et al. 2021). Due to its relatively slow kinetics, this system is more suitable for optical manipulation of lower frequency stimulation. Additionally, optical activation of these opsins, which have a long opening lifetime, may elevate the concentration of intracellular ions with unintended physiological complications (see Discussion).

For these reasons, we created a system with fast closing kinetics. We chose vfChrimsoncurrently the fastest R-ChR-to gain precise temporal control of red-light excitation (Mager et al. 2018). To generate a B-ACR with vfChrimson matching kinetics we used a model-based mutagenesis approach to modify the kinetic properties of ZipACR, an ultrafast blue light-activated chloride channel (Govorunova et al. 2017). It must be noted that due to the high intracellular chloride concentration in the axonal terminal (Turecek and Trussell 2001; Messier et al. 2018) the use of a chloride channel-based expression system must be restricted to the somato-dendritic regions (Mahn et al. 2016; Messier et al. 2018).

We describe several variants of ZipACR, focusing on two, I151T and I151V, which display appropriate kinetics, light spectra, and photocurrent magnitudes for the purposes of our co-expressing system. We then paired the optimized anion channels with a membrane-trafficking optimized vfChrimson (IvfChr) in a single expression unit. The new system, Zip-IvfChr, when expressed in neurons, can drive time-locked high-frequency action potentials (APs) in response to red pulses (635 nm). On the other hand, in response to blue pulses (470 nm), Zip-Ivf-Chr produces no action potentials.

As expected, blue light pulses transiently suppress APs in Zip-IvfChr-expressing neurons, but the suppression is fully reversed within 5 ms after termination of light pulses due to the use of fast ZipACR mutants. When tested in the facial motor nucleus in the brainstem, Zip-IvfChr activation by red but not blue light triggered vibrissa movement of large amplitudes. Due to its high temporal precision of activation and suppression, this tool could be further developed for independent optical excitation of distinct neural populations.

Results

The strategy to reduce blue light-mediated excitation by red-shifted channelrhodopsins

We initially tested our strategy using existing variants of excitatory and inhibitory channelrhodopsin. The channelrhodopsin Chrimson and its variants have the most red-shifted spectra (Klapoetke et al., 2014), and their response to 470 nm has a slower on-rate compared to the 590 nm light (Figure 1-figure supplement 1A and B). Despite this, 470 nm light is equally effective in inducing APs (Figure 1-figure supplement 1C). To overcome this limitation, we decided to take advantage of blue light-gated inhibitory chloride channels which effectively block action potential upon exposure to blue light (Govorunova et al. 2015, 2017). By co-expressing the B-ACR variant GtACR2, along with a Chrimson variant, photostimulation of targeted neurons by red light would be expected to only activate R-ChR, which causes photoactivation of these neurons. Blue light, however, activates R-ChR as well as GtACR2. This allows the chloride channel to induce shunting inhibition of the neurons which in turn would be expected to prevent firing of the neurons by blue light (Fig. 1A).

Reducing blue-light mediated excitation in red-shifted channelrhodopsins- Proof of concept.

(A) Schematic representation of the approach. A red-shifted channelrhodopsin (R-ChR) is co-expressed with a blue-shifted anion channel (B-ACR). When red light is ON, there is an overall excitation of the cell due to the more dominant R-ChR response. When blue-light is ON, the shunting inhibitory effect of B-ACR reduces the excitation induced by R-ChR to blue light. This panel was created using BioRender.com. (B) Validation of the approach in flies expressing Chrimson, Chrimson and GtACR2, or GtACR2 alone. (B1) Experimental setup to record the courtship song of solitary male flies. This panel was created using BioRender.com. (B2) Top, Example of a reconstructed neuronal arborization of P1 neurons, Bottom, example of a typical courtship song induced by LED pulses. (B3) Courtship song production of the solitary male flies during 10-sec of constant illumination with 590 nm (amber) or 470 nm (blue) light. Each dot represents an individual drosophila (mean ± SEM).

Proof of concept of the co-expression strategy in Drosophila

As the first step to test the potential validity of this approach, we took advantage of the well-studied male-specific P1 neurons in the fly central brain that elicit courtship singing upon activation (von Philipsborn et al. 2011; Inagaki et al. 2014; Ellendersen and von Philipsborn 2017; Mohammad et al. 2017) (Fig. 1B1-B2). When Chrimson was co-expressed with an inactive control protein (tdTomato), illumination at 470 nm and 590 nm wavelengths led to similar amounts of courtship song (110.1 ± 12.8 and 78.5 ± 11.6 pulses per 10s, n = 16-17, respectively), demonstrating that blue light at the chosen intensity can efficiently activate Chrimson (Fig 1B3). When Chrimson was co-expressed with the blue-sensitive anion channelrhodopsin GtACR2, 590 nm light illumination reliably elicited male courtship song (95.2 ± 9.14 pulses per 10s); while 470 nm light illumination did not evoke any detectable pulse song (n = 17-18). Neither 590 nm nor 470 nm light pulses exposure induced courtship song in flies expressing only GtACR2 (n = 5).

Identifying a suitable pair of ultrafast light-gated anion channel and a red-shifted Chrimson variant

Although GtACR2 proved to be effective in protocols with pulses of long duration, its slow current decay (Fig. 2A, see GtACR2, 161 ± 24 ms, n = 3) limits its application to optical stimulations of lower frequencies (Govorunova et al. 2015, 2017). For this reason, we searched for a pair of blue-shifted anion channelrhodopsin and red-shifted cation channelrhodopsin with fast and matching kinetics. For a fast R-ChR, we evaluated the recently published vfChrimson, the fastest red-shifted channelrhodopsin available at this time, which can induce APs up to the intrinsic limit of the neurons (Mager et al. 2018).

Comparison of channel properties of light-gated chloride channels and Chrimson variants.

(A) Representative photocurrents of the Chrimson and ACRs variants in whole-cell patch-clamp recordings from HEK 293 cells with 590 nm and 470 nm illumination (top black bars, 10 mW/mm2). The plot shows the off-time constant (top) following 1 sec of 470 nm illumination, and the peak photocurrent (bottom) at -60 mV (mean ± SD, n = 3-13 cells). (B) Typical response of CA1 cells from hippocampal organotypic slices expressing ZipACR to 40 Hz, 5 ms pulses (top black trace) and to an illumination protocol using 40 Hz of 10 ms light pulses at 470 nm (4 mW/mm2) (bottom blue trace). (C) Recordings from dentate gyrus granule cells in acute slices expressing ZipACR and vfChrimson. (C1) Representative responses to 500 ms illumination (bottom black bar) at 470 nm and 635 nm (10mW/mm2) (blue and red traces respectively). Note, 635 nm illumination evoked action potential firing while 470 nm illumination only evoked a single action potential time-locked to the end of the light pulse (arrowhead). Insets, example of an action potential upon 470 nm light termination and plot showing the number of light-evoked spikes at 635 nm (red bar) and 470 nm (blue bar) illumination. n = 10 cells, paired t-test **p< 0.01; (C2) Example of light-induced firings by 635 nm and 470 nm illumination (red and blue traces respectively, 5 ms pulses (black squares) at 10 Hz). The arrowheads show consistent APs time-locked to the end of the blue light pulse. Inset, example of a voltage-clamp recording of currents evoked by 5 ms pulses of 635 nm and 470 nm light (red and blue traces respectively). Light-induced currents with 470 nm illumination were initially outward but turned inward gradually and peaked immediately following the light offset. Note that the late inward current matches the red light induced current, reflecting a slower kinetics for vfChrimson compare to ZipACR.

Previously, we found that Chrimson and its variants traffic poorly to the membrane and developed a strategy to improve the trafficking (C-Chrimson-ts, (Bonaventura et al. 2019). We adapted this modification to improve the trafficking of vfChrimson (referred to as IvfChr, Figure 2 - figure supplement 1). In our measurements in HEK 293 cells, IvfChr has a channel off-rate consistent with previously reported (Mager et al. 2018) (off-rate time constant of 5.6 ± 0.3 ms, n = 10 compared to C-Chrimson-ts 21.5 ± 1.2, n = 10; mean ± SEM, Fig. 2A, Figure 2 - figure supplement 2). When comparing C-ChrimsonR-ts and IvfChr photocurrents, we observe 40% reduction of photocurrent of IvfChr (0.086 ± 0.055 pA/ pF/A.U, n = 25) compared to C-ChrimsonR-ts 0.145 ± 0.126 pA/pF/A.U., n = 25; mean ± SD, Figure 2 - figure supplement 2B) when the peak photocurrent amplitude was adjusted to membrane capacitance and membrane fluorescence.

To counteract the IvfChr excitation, we estimated that desired properties of the blue light activated anion ChR should have the off-rate time constant of ∼15-30 ms, minimal response to light of wavelength > 590 nm and a strong photocurrent. We identified ZipACR, RsACR_995, GcACR_457, RlACR_477 and GtACR2 as possible templates for further development based on published data (Govorunova et al. 2017). GtACR2, RsACR_995 and RIACR_477 are spectrally ideal with negligible response to 590 nm light (Figure 3 - figure supplement 1A), but their kinetics are slow (mean off-rate of 161.0 ± 24.1, 122.7 ± 36.3 and 95.8 ± 49.4 ms, respectively, n = 3, Fig. 2A) which could have undesired long lasting inhibitory effects after light termination. GcACR_457 has very small photocurrents and cannot be characterized accurately, possibly due to poor membrane trafficking. ZipACR, although spectrally not ideal (∼515 nm; (Govorunova et al. 2017) has a kinetic close to the desired range (9.5 ± 3.0 ms, n = 9) and it has strong photocurrents (Fig. 2A).

Characterization of the optimized variants of the ultrafast anion channelrhodopsin ZipACR.

(A) Structure-based design of ZipACR variants. (A1) The retinal binding pocket of GtACR1 and the putative retinal binding pocket of ZipACR homology model. (A2) The alignment of the transmembrane domains (TM) of GtACR1, GtACR2 and ZipACR. The residues surrounding the retinal binding pocket of GtACR1 and the corresponding residues from GtACR2 and ZipACR are highlighted in red. *marks the residues targeted for mutation. The lysine that forms the Schiff base with retinal is indicated in green. The numbers on the left correspond to the number of the first residue in each line. (B) Basic properties of the ZipACR variants compared to IvfChr. Off-time constant (top) following 1 sec of 470 nm illumination, peak photocurrent (middle) and 590/470 ratio (bottom). The data have been obtained from whole-cell recordings in HEK 293 cells. Each dot corresponds to one cell (mean ± SD, n = 2-12). (C) On-time constant (C1) and normalized photocurrent (C2) of the selected variants Zip(151T) and Zip(151V) as compared to the original ZipACR and IvfChr, with 470 nm or 590 nm illumination of various light intensities (in mW/mm2). For the photocurrent, the values have been measured at the maximum response. The data points are the mean value ± SEM (n = 6-7). (D) The responses of the selected variants to various wavelengths of the same photon flux. (D1) The action spectra of ZipACR, Zip(151V) and Zip(151T) as measured from 650 nm to 400 nm (D2) For the action spectra, the maximum photocurrent amplitudes measured at each wavelength were normalized to the peak values obtained from the same cell across the spectrum. The data points are the mean value ± SEM (n = 6-7). (E) Validation of the efficiency of the Zip(151T) variant in slices. (E1) The response of dentate gyrus granule cells expressing Zip(151T) variant at 470 nm (4mW/ mm2) and 635 nm (7 mW/mm2) light pulses delivered through a 63x objective. This panel was created using BioRender.com. (E2) Representative traces of the firing induced by current injection at 40 Hz. Overlapping 470 nm but not 635 nm pulses block action potentials. The plot on the right shows the efficiency (in %) of Zip(151T) in blocking individual action potentials at 470 and 635 nm light pulses (mean ± SD, n = 6-10 cells). The traces of the two outliers are presented in Figure 3 - figure supplement 3A.

To test the inhibitory property of ZipACR, we injected 40 Hz somatic current in organotypic hippocampal neurons expressing the channel. Consistent with the original report, individual pulses of 470 nm light resulted in the time-locked suppression of APs (Fig. 2B). Next, we examined whether ZipACR can block blue light-mediated APs in neurons expressing vfChrimson. Recordings from the granule cells in acute hippocampal slices co-expressing ZipACR and vfChrimson showed that, in 8 out of ten cases 470 nm light induced significantly fewer APs than 635 nm light (Fig. 2C1), in line with our prediction. However, we consistently observed the appearance of individual APs time-locked to the blue light offset (Fig. 2C1-C2). We concluded that due to the fast-closing rate, the photoinhibitory current generated by ZipACR decays too early to suppress the remaining excitatory response of vfChrimson. Therefore, we decided to generate variants of ZipACR with slightly slower off-kinetics.

Optimizing ZipACR off-kinetics

It is technically easier to generate mutations with slower kinetics and blue-shifted spectra than generating variants with faster on- and off-kinetics. Thus, we decided to use ZipACR as a template for further mutagenesis instead of further modifying GtACR2. To identify relevant mutation sites in ZipACR, we generated a homology model of ZipACR with the crystal structure of GtACR1 (Govorunova et al., 2017) and identified the putative residues forming the retinal binding pocket in ZipACR (Fig. 3A). Multiple alignments were performed with the kinetically slower GtACR1 and GtACR2 to identify possible replacement of these residues that would result in slower kinetics and/or small blue-shift in ZipACR without disrupting the channel function (Fig. 3A2). We tested these variants in HEK 293 cells for photocurrent ratio 590 nm / 470 nm LED stimulating light and measured the channel off-rate kinetics and photocurrent amplitude (Fig. 3B).

Of these candidates, we identified I151V and I151T at the 4th transmembrane domain of ZipACR that resulted in a small increase in channel off-rate time constant (15.5 ± 8.4 ms for I151V and 20.2 ± 6.4 ms for I151T, mean ± SD, Fig. 3B) compared to the original ZipACR (9.5 ± 2.9 ms, mean ± SD). We also generated double and triple mutants combining the I151V or I151T with Y170F, Y177F or V229L mutations but most of these double or triple mutants appeared to have smaller photocurrents and off-rate kinetics that could not be fitted well with single exponential back to zero baseline. Based on these results, we concluded the variants I151V (Zip(151V)) and I151T (Zip(151T)) are the best candidates for the pairing and proceeded for further characterizations.

Neither Zip(151V) nor Zip(151T) mutations changed the ion permeability of ZipACR as measured by channel reversal potential in potassium gluconate and cesium chloride based intracellular recording solutions (Figure 3 - figure supplement 1B and C). In terms of light sensitivity, it is desirable that the ZipACR variant would respond to 470 nm light faster and stronger and to red-shifted light slower and weaker than IvfChr at the same intensity (Figure 3 - figure supplement 2). With both Zip(151V) and Zip(151T), the channel on-rate kinetics have the desirable properties as described above to 470 nm and 590 nm light, although the Zip(151T) on-rate starts to approach the on-rate of IvfChr at light intensities >5 mW/mm2 at 590 nm (Fig. 3C1). With 470 nm light, IvfChr does not appear to reach saturating response even at 20 mW/mm2 (Fig. 3C2). With 590 nm light, IvfChr reaches saturating response at 10 mW/mm2 whereas none of the ZipACR variants appears to reach saturating response even at 20 mW/mm2 (Fig. 3C2). Spectrally, Zip(151T) shows a wider red-shifted action spectra than ZipACR whereas Zip(151V) may have a slightly narrower blue-shifted action spectra than ZipACR (Fig. 3D).

Since Zip(151T) is a more red-shifted variant of the two (Fig. 3B,D), we asked whether red light pulses may induce photoinhibitory currents that suppress APs. We expressed Zip(151T) in the granule cells of the dentate gyrus in hippocampal slices using AAV vector and tested its inhibitory effect upon 470 nm and 635 nm light pulses. We injected somatic current at 40 Hz, while alternately delivering overlapping 5 ms pulses of light. Except in two instances, blue light pulses were consistently effective in blocking APs, while red light had no effect on APs induction (Fig. 3E, n = 10 for 470 and n = 6 for 635, the two outliers are presented in Figure 3 - figure supplement 3A). It has been reported that at resting or hyperpolarized membrane potentials ZipACR may produce light-evoked APs (Kato et al. 2018). We did not observe such a phenomenon under our experimental setup (Figure 3 - figure supplement 3B).

Validation of the co-expression strategy in brain slices

Since co-injection of two viruses results in variability in the ratio of expression in different neurons, we used the bicistronic 2A approach (Supplementary material 1). We favored this system to the internal ribosomal entry sites (IRES) (Pelletier and Sonenberg 1988; Douin et al. 2004) and tandem fusion strategies (Kleinlogel et al. 2011). As observed by us and reported by other groups, IRES produces an unbalanced expression level with the second protein at much lower level (Mizuguchi et al. 2000). The testing of vfChrimson in the recently published BiPOLES tandem fusion system with GtACR2 (Vierock et al. 2021) resulted in high levels of intracellular accumulation of the tandem opsins in HEK293 cells (Figure 2 - figure supplement 1).

Therefore, we generated constructs containing IvfChr and Zip(151T/V) (ZipT-IvfChr and ZipV-IvfChr) using the bicistronic 2A cassette with KV2.1 soma targeting sequence for both ZipACR mutant and IvfChr (Fig. 4A, Figure 2 - figure supplement 1 and Supplementary material 1). This ensures an equimolar translation of the two proteins from one mRNA and soma enrichment of the opsins in neurons. In comparison to BiPOLES with vfChrimson, the vfChrimson in our bicistronic 2A cassette led to ∼2.7 x higher photocurrent (−7.57 ± 5.37 pA/pF in ZipV-IvfChr compared to -2.72 ± 1.33 pA/ pF in BiPOLES, n = 14) when tested in identical conditions, consistent with the improved membrane trafficking observed with IvfChr (Figure 2 - figure supplement 1A and B). The photocurrent amplitudes of the IvfChr and ZipV-IvfChr constructs in HEK 293 cells are comparable after adjustment to cell size and membrane fluorescence (Figure 2 - figure supplement 1D). We did not test the ZipT-IvfChr variant as the red-shifted response of Zip(151T) makes it more difficult to isolate the IvfChr response to 590 nm light available on our LED system.

Zip(151T) variant co-expressed with the red ChR2 IvfChr prevents blue-light induced action potentials, in-vitro and in-vivo.

(A) The expression cassette for IvfChr and bicistronic ZipT-IvfChr used in rAAV vector. The selected ZipACR variant Zip(151T) is co-expressed with the red-sensitive IvfChr by using the 2A self-cleaving peptide (T2A). All the opsins are soma-targeted with a Kv2.1 peptide. (B) Comparison of the high-frequency red and blue light-induced spiking (B1), and spike probability (B2, n = 6-13 for each opsin), of DG cells expressing IvfChr or the ZipT-IvfChr, with 10 and 20 Hz of light pulses (4-7 mW/mm2). Holding potential = -60 mV. (C) Blue and red lightdriven spike fidelity at various light intensities in DG cells (n = 4-10 cells for each opsin; 10 ms pulse width, single pulse). (D) Experimental design. ZipT-IvfChr or IvfChr are expressed in the facial nucleus (FN) of the brainstem (top) and light-induced whisker movements are recorded. Example of soma-targeted expression of our ZipT-IvfChr construct in FN (bottom). This panel was created using BioRender.com. (E) Averaged Z-scores of the angle of the whiskers under 450 nm and 638 nm illumination (20 mW/ mm2) of the FN motoneurons. ZipT-IvfChr does not trigger whiskers movement under blue-light (n=4 animals per condition). (F) Ratio of the AUC of the responses to 450 nm and 638 nm illumination for ZipT-IvfChr and IvfChr (n=4 animals per condition, **p < 0.01, t-test). (G) Distribution of the AUC for three individual trials from each animal at 450 nm and 638 nm illumination (n=4 animals per condition).

After characterization in HEK293 cells, we examined the power at which responses to 470 nm and 635 nm lights induce APs in neurons expressing ZipT-IvfChr, ZipV-IvfChr, or IvfChr. We expressed AAV constructs in the granule cells of the dentate gyrus in the hippocampal slices (as described in Fig. 3E1). We recorded cells under the current-clamp mode and injected current through the patch pipette to maintain the membrane potentials at −60 mV. To examine the efficiency of 635 nm and 470 nm lights in inducing APs, we delivered 10 Hz and 20 Hz light pulses of 5 ms duration (Fig. 4B). As expected, in IvfChr expressing neurons, both 470 nm and 635 nm lights induced APs (n=6). In neurons expressing ZipT-IvfChr (Fig. 4B), 635 nm pulses were equally effective in driving APs. On the other hand, 470 nm light only induced modest depolarization, which did not produce the threshold for AP induction (n = 5-12 for each opsin). Similarly, co-illumination with 470 nm and 635 nm also failed to generate AP, showing blue light effectively nullifies the depolarizing property of red light (Figure 4 - figure supplement 1A).

In IvfChr expressing neurons 470 nm and 635 nm lights induced APs with light powers as low as 1 mW/mm2 (n = 10, Fig. 4C). In contrast, ZipT-IvfChr expressing neurons produced APs only to 635 nm light stimulation (Figure 4 - figure supplement 1B), whereas 470 nm light powers as high as 10 mW/mm2 failed to induce APs (n = 7-8, Fig.4C). As for ZipV-IvfChr, we occasionally observed blue light-induced APs (Figure 4 - figure supplement 1B), which may be explained by its relatively faster closure rate (Fig. 3B). In neurons expressing IvfChr, we noticed a reduction in probability of AP upon successive optical stimulation specifically to 635 nm pulses (Fig. 4B). Consistent with this, we observed a reduction in response in HEK 293 cells expressing IvfChr (Figure 4 - figure supplement 2). This phenomenon, likely caused by the desensitization of the channel, is consistent with previous reports with other red-shifted channelrhodopsins (Lin et al. 2013).

Validation of the co-expression strategy in vivo

To evaluate the system in vivo, we compared the performance of ZipT-IvfChr and IvfChr in driving vibrissae movements by optically activating motoneurons of the facial nucleus (FN) expressing either of the two constructs (Lin et al. 2013; Sreenivasan et al. 2015; Gharaei et al. 2020) (Fig. 4D). Four to six weeks after virus injection, a 200-um diameter optic fiber was placed above the FN of head-fixed anesthetized mice and motoneurons were stimulated at 0.33 Hz using 100ms pulses of either 450 nm or 638 nm light. In mice expressing IvfChr, whiskers movements of large amplitude were elicited with both red and blue pulses (Fig. 4E-G, the Area Under the Curve (AUC) of 2.9 ± 0.09 and 1.95 ± 0,13, respectively, with a ratio AUC450/AUC638 = 0.68 ± 0.05, n=4) (Supplementary material 2). Pulses of red light elicited strong whisking (AUC of 1.8 ± 0.29, n=4) in mice expressing ZipT-IvfChr, which was comparable to the IvfChr group. On the other hand, 450 nm light pulses failed to reliably elicit whiskers movement, as represented by a significant decrease of the AUC under blue light illumination (0.25 ± 0.07, ratio AUC450/AUC638 = 0.15 ± 0.04, n=4, p= 0.0012, t-test) (Supplementary material 2). Overall, we confirmed that in vivo, unlike motoneurons expressing IvfChr, motoneurons expressing ZipT-IvfChr are largely non-responsive to blue light, while preserving the red shifted property of an R-ChR.

Fast recovery of blue-light induced suppression of action potentials in vitro

We expected that blue-light pulses in ZipV-IvfChr- and ZipT-IvfChr-expressing cells would have an overall inhibitory effect on membrane potential due to the activa-tion of ZipV/T but this effect is short in duration due to the fast kinetics of ZipACR variants. To test this, we stimulated the neurons by somatic current injection at 40 Hz, while alternately delivering perfectly overlapping 5 ms pulses of lights (Fig. 5A). Light pulses of 470 nm blocked APs, while 635 nm light pulses had virtually no effect. Importantly, we asked how fast a neuron recovers from the blue-light inhibition associated with the expression of ZipV-IvfChr and ZipT-IvfChr. We induced APs through somatic current injection at different time points from the offset of a light pulse (Fig. 5B1). In neurons expressing ZipT-IvfChr, current injection evokes APs with high probability after 5 ms, which is consistent with our goal of minimizing the unintended disruption of membrane excitability after the light stimulation (Fig. 5B). The recovery time for ZipV-IvfChr was immediate, with 100% success rate immediately following the light offset (Fig. 5B2).

The blue-light induced suppression of action potentials is confined within the duration of the light pulse.

(A) Optical inhibition of spiking in DG cells expressing the ZipT-IvfChr and ZipV-IvfChr variants. Representative traces of the firing induced by current injection at 40 Hz. Overlapping 470 nm but not 635 nm light pulses block action potentials. Holding potential = -60 mV. Plots on the right shows the efficiency (in %) of ZipT-IvfChr in blocking individual action potentials at 470 and 635 nm (n = 9-10). (B) Temporal recovery of neuronal excitability following light-induced inhibition by the ZipACR variants. (B1) Experimental paradigm. DG granule cells expressing ZipT-IvfChr or ZipV-IvfChr have been stimulated by a somatic current injection (5 ms pulse) overlapping or preceded by a 470 nm pulse (5 ms, 4 mW/mm2) at 0, 5, 10, 15 or 20 ms intervals. The traces in current clamp (bottom) are representative of the responses to current injections. This panel was created using BioRender.com. (B2) The spike probability shows a fast recovery of ZipT-IvfChr (<5ms) and an immediate recovery of ZipV-IvfChr, from the off-set of the 470 nm light (n = 7 for ZipT-IvfChr, n = 6 for ZipV-IvfChr, mean ± SD).

Evaluation and modification of recently developed light-gated potassium channelrhodopsins

A main drawback of using chloride channels is that their opening in the axonal terminals causes depolarization and neurotransmitter release (Turecek and Trussell 2001; Mahn et al. 2016). This is due to the high concentration of chloride in the terminals. Since potassium channels are hyperpolarizing in all neuronal compartments, we evaluated the recently published potassium selective channelrhodopsins HcKCR1, HcKCR2 (Govorunova et al. 2022), WiChR and B1ChR2 (Vierock et al. 2022) for possible pairing with IvfChrimson in next version of our tool. We observed significant light-induced membrane currents from HEK293 cells in all the examined potassium-selective channelrhodopsins, except for B1ChR2 (Figure 5 - figure supplement 1A) This is consistent with previously reported data in mammalian cells (Vierock et al., 2022). Therefore, B1ChR2 was not tested any further.

The channel reversal potentials, as measured by a ramp protocol, were -58.0 ± 8.7 mV (n = 9), -50.7 ± 8.0 mV (n = 14) and -65.2 ± 6.5 mV (n = 7) for HcKCR1, HcKCR2 and WiChR, respectively (Figure 5 - figure supplement 1B and C) which were more hyperpolarized than the reversal potential for ZipACR measured previously in similar conditions (−39.0 mV). In terms of the channel kinetics, HcKCR1, HcKCR2 and WiChR have the off-rate time constants (measured at holding potential of -10 mV) of 49.7 ± 11.3 ms (n = 9), 58.6 ± 47.8 ms (n = 14) and 229.3 ± 53.8 ms (n = 9) (Figure 5 - figure supplement 1D) which is much slower than the channel kinetics of ZipACR.

As an observation, the off-rate for the HcKCRs often do not fit a single exponential function well and the value measured here underestimate the speed of the channel (see the traces in Figure 5 - figure supplement 1E1). HcKCR1 has spectral peak at 530 nm as previously reported (Govorunova et al. 2022) and still responds strongly to 590 nm light (73.0 +/-6.4 % of 530 nm response, n = 10) whereas HcKCR2 and WiChR have spectral peaks at 470 nm and have small responses to 590 nm light (9.0 +/-3.0 % of 470 nm response, n = 8 for HcKCR2 and 27.6 +/-7.2 % of 470 nm response, n = 6; for WiChR; Figure 5 - figure supplement 1E2). In light of these results, none of the wild-type K-ChRs appear to be optimal for the pairing and further mutagenesis, and engineering in future study would be required to optimize the pairing.

Discussion

Here, we describe a dual-color expression system by which pulses of red-light drive time-locked high frequency APs, while blue light allows precise suppression of AP confined within the duration of the pulse. We created this system by pairing a B-ACR with the ultrafast red-shifted ChR, IvfChr. This requires the B-ACR to have a significantly larger photocurrent compared to R-ChR as well as a faster channel opening and slightly slower channel closing. As we discuss later, we considered anion channels with slow decay of inhibition are not suited for many experimental manipulations of long duration. For this reason, we chose ZipACR, currently the fastest B-ACR, as the template for further modifications. Based on our homology modeling and multiple alignments with the kinetically slower B-ACRs, GtACR1 and GtACR2, we identified potential residue replacements within the putative retinal binding pocket which would result in slightly slower kinetics without impairing the channel function.

To our surprise, many mutations of ZipACR around the alltrans retinal do not alter the channel properties or compromise channel conductance very strongly. This is contrary to our past experiences with channelrhodopsins where conductance is highly sensitive to mutations around the retinal binding pocket (data not shown). ZipACR double mutants appear to have more compromised conductance or altered properties. Among the tested positions, the I151 residue appears to have significant influences in kinetics and spectral properties. Indeed, we found that two variants, Zip(151T) and Zip(151V), have properties optimal for our purpose. More specifically, the channels off-rate kinetics have the desirable property of being slightly but sufficiently slower than IvfChr without compromising the opening rate, photocurrent and light spectrum. Thus, when paired with IvfChr, they prevent AP induction by blue light, without interfering with red light-induced high frequency neural spiking.

Of the two variants, we observed Zip(151T), and consequently ZipT-IvfChr, to be more consistent in blocking blue light-induced APs, which could be explained by its slightly slower closing rate. Although Zip(151T) is the more red-shifted of the two, within the tested range of light intensities and expression, only rarely did we observe a block of APs at 635 nm. However, we do not recommend the use of orange light pulses with ZipT-IvfChr, as we observed a significant photocurrent in this wavelength (Fig. 3D1 and D2; Figure 3 - figure supplement 2). Comparing the ZipV-IvfChr and the ZipT-IvfChr constructs, we would recommend users to use the ZipT-IvfChr for most general experiments if fast light source with wavelength > 600 nm is available given the reduced probability of neuronal spiking at the end of blue light pulse. The ZipV-IvfChr system is faster in kinetics and is suitable in situation where kinetics is critical but there is some tolerance in spiking at the end of blue light pulse in some neurons.

A recent study (Vierock et al. 2021) proposed an R-ChR expression system by relying on the pairing between GtACR2 and Chrimson, as we did in our early stage work with Drosophila. For experimental reasons linked to controlling neuronal physiology, the priority we set was to create a system with fast closing kinetics to minimize any physiological disruption after the termination of light illumination. All channelrhodopsins have the potentials to alter osmotic balance, ion gradient and pH of small neuronal compartments and this is exacerbated by long light pulses and prolonged channel opening. This can be problematic in long-term neuroplasticity experiments. Cationic channelrhodopsins are known to conduct calcium ions, long opening can evoke significant elevation of intracellular calcium (Lin et al. 2013) and potentially activate second messenger pathways.

The alteration of pH, osmotic balance and ion gradient (either directly or secondary to elevated transporter activity) applies to both cationic and anionic channelrhodopsins and can have unpredictable biological consequences. Hence, channelrhodopsins with rapid kinetics such as the ones used in our system and brief stimulation light pulses are more ideal for experiments where neuronal or synaptic plasticity are of importance. We were able to demonstrate that neurons expressing ZipT-IvfChr can recover from this inhibition within 5 ms after light termination and ZipV-IvfChr recovers almost instantaneously. If we model our results to a B-ACR with slower kinetics such as GtACR2 used in BiPOLES, we could expect the inhibitory effects of blue light to last up to 10-20 ms. This could disrupt normal neuronal physiology and signal processing that occurs in millisecond time scale, as well as alter intracellular ion concentrations which may lead to short-term and long-term alteration in physiology.

From an experimental viewpoint, a fast photostimulation system has the flexibility to perform a wide range of optical manipulations. Ultimately, ZipT-IvfChr and ZipV-IvfChr will allow dual optical activation of segregated groups of neurons when co-expressed with a fast blue-shifted channelrhodopsin such as oChIEF in the same brain region. In these experiments, it is critical to minimize the blue-light mediated inhibition due to possible disruption of responses resulting from local monosynaptic connection that may have important modulatory effects on postsynaptic cells. Spike-timing dependent experiments requiring high temporal precision would be one of the experiments that would benefit from our system. Blue-light activated cationic channelrhodopsin such as oChIEF can be expressed presynaptically to control neurotransmitter release and ZipV/T-IvfChr can be expressed postsynaptically to control spiking to induce long-term potentiation or long-term depression.

In cortical experiments, blue cationic channelrhodopsin and ZipV/T-IvChr can be expressed in projection neurons and interneurons to study the effects of excitation/inhibition balance in cortical processing of information or decision making. In the striatum, the independent control of neurons of direct and indirect pathways would benefit from the dual wavelength manipulation. Many of these experiments would require the use of retrograde AAV, cell-type specific enhancer/promoter or transgenic animals with defined driver line. Although ZipV/T-IvfChr can theoretically be used with single-photon jGCaMP imaging, this would require blue light illumination with duration greater than several milliseconds which will lead to prolonged inhibition of ZipV/T-IvfChr-expressing.

The main limitation of our current approach is that there is limited control of the IvfChr and ZipV/T expression. Although the 2A approach should achieve better translation ratio compared to the use of Internal Ribosome Entry Site (IRES), the levels of respective proteins would also depend on the amount of protein degradation and the fraction of channelrhodopsins that failed to fold or incorporate the retinal chromophore correctly. This can vary with different neurons in different regions. Therefore, similar to other optogenetic systems, it is important for the users of our system to validate the approach in their cells of interest before commencing behavioral experiments.

There can also be situations where dual virus delivery would be preferable compared to bicistronic expression cassette. As shown in our testing, the tethered-tandem expression system used in the BiPOLES reduces the photocurrent of vfChrimson compared to our IvfChr, possibly associated with hindered membrane trafficking. Given the reduced conductance of vfChrimson compared to Chrimson, a BiPOLES based tethered-tandem expression system is also not ideal. The development of an improved tunable dual expression system would therefore be beneficial for the neuroscientific community. Ultimately, the co-expression strategy remains a temporary solution to the spectral cross-talk of opsins and the development of a true spectrally narrow channelrhodopsin would still be the preferred solution. However, whether this is achievable due to the chemical properties of the retinal chromophore is questionable.

Before concluding, we must note that our approach, which acts based on chloride flux, is unlikely to be suitable for axonal terminal inactivation, where the intracellular chloride concentration is high (Mahn et al. 2016, 2018). Potassium selective channelrhodopsins, having a hyperpolarized reversal potential in all cellular compartments, may appear better positioned for our purpose. However, as we show in our characterizations, none of the wild-type variants would be optimal for the pairing kinetically. WiChR is kinetically too slow but is the most potent inhibitor. HcKCR1 has strong desensitization, is kinetically too slow and spectrally too red-shifted. HcKCR2 is the least potassium selective but is spectrally and kinetically closest to the ideal candidate of the three. Overall, further engineering and modification would still be required to identify the ideal mutants for the pairing similar to what we have done with ZipACR in this study.

At present, for axonal or terminal stimulation, the prior published approaches utilizing differential kinetics and light sensitivity of the opsins or optical manipulation combined with synaptic release properties would still be the preferred options (Klapoetke et al. 2014; Hooks et al. 2015). Specifically, we refer our readers to a recent work, where we demonstrate effective independent dual optical activation of converging axons, electrophysiologically as well as behaviorally (Faress et al. 2024). Thus, the system introduced here complements these studies that are unsuitable for somatic photoactivation. This permits researchers to choose complementary methods based on their experimental needs. Future engineering of a faster and blue-shifted potassium selective channelrhodopsin would provide and unlock a more comprehensive approach suitable for both somatic and axonal stimulation (Govorunova et al. 2022; Vierock et al. 2022).

Materials and Methods

Drosophila optogenetic activation experiments

Flies (Drosophila melanogaster) were kept on standard food medium (water, cornmeal, oatmeal, sucrose, yeast, agar, acetic acid, preservative methyl-4-hydroxybenzoate) at 25°C, 60% humidity, and a 12h/12h light-dark cycle. Male flies were collected after eclosion and aged for 7 days in isolation and protected from light on standard food, followed by 3 days on food containing 400 µM all-trans retinal (Sigma). The following genotypes were assayed: co-expression of Chrimson and GtACR2: w-; NP2631.GAL4/ UAS-CsChrimson.mVenus. attp40; tubP-FRT-stop-FRT-GAL80, fruFLP/ UAS-GtACR2. attp2; co-expression of Chrimson and tdTomato: w-; NP2631. GAL4/ UAS-CsChrimson.mVenus.attp40; tubP-FRT-stop-FRT-GAL80, fruFLP/ UAS-myr-tdTomato. attp2; co-expression of GtACR2 and tdTomato: w-; NP2631. GAL4/ UAS-myr-tdTomato.attp2; tubP-FRT-stop-FRT-GAL80, fru-FLP/ UAS-GtACR2.attp2.

For optogenetic activation experiments, we used a previously described multi-channel array of electret condenser microphones (CMP-5247TF-K, CUI Inc), amplified with a custom-made circuit board and digitized with a multifunction data acquisition device (NI USB-6259 MASS Term, National Instruments)(“Cellular and Behavioral Functions of Fruitless Isoforms in Drosophila Courtship” 2014) (von Philipsborn et al. 2014). Single male flies were placed in recording chambers equipped with microphones at the bottom and light diodes (Amber 591 nm: Cree 5-mm C503B-AAS/AAN-015 or Blue 470 nm: Cree 5-mm C503B-BAS-CY0C0461) were placed 1 cm above the song recording chambers at a 60° angle, illuminating the whole chamber (10-sec of constant light). Tentative courtship song was detected by a MATLAB script and corrected manually, using a custom made user interface for visualization and annotation of song oscillograms (von Philipsborn et al., 2014).

Cell culture, transfection and stable cell line generation

HEK 293 cells (Invitrogen) were grown in DMEM (1g/L glucose) supplemented with 8% FBS and 1% Pen/Strep. For recordings, cells were plated at low density on glass coverslips and kept in DMEM media supplemented with 4% FBS and 1% Pen/Strep. Transfection was done with Xtremegene 9 (Sigma-Aldrich) according to the manufacturer’s instructions. Recordings of the cells were performed between 38-50 hours after transfection.

For generation of recombinant lentivirus for stable HEK 293 cell lines expressing IvfChr and ChrimsonR-tdTomato cells, HEK 293 cells were co-transfected with psPAX2 and pMD2.G (courtesy of Trono laboratory) and the media was transferred to a new dish with untreated cells 2 and 3 days after transfection. The cells receiving the media were kept as stable cell lines of respective constructs.

Structure alignment and modeling

Homology modeling of ZipACR structure was performed with Modeller within the UCSF Chimera to the published GtACR1 structure (PDB #6CSM). This was further aligned to the GtACR2 alignment from (Govorunova et al. 2017). Residues surrounding the retinal binding pocket of GtACR1 and the modeled ZipACR were identified and those residues that were different in ZipACR were changed to the corresponding residues of GtACR2 or GtACR1.

Molecular biology

For GcACR_457, RsACR_995 and RlACR_477, the cDNA were in the original pcDNA3.1 vector as EYFP fusion as available on Addgene (#103137, 103772, 103771). The GtACR2 construct was synthesized (IDT DNA) and modified with a N-terminal signalling peptide from the commercial pDisplay vector (Invitrogen) and a trafficking signal prior to tdTomato or citrine fusion and inserted in the pcDNA3.1+ vector (Invitrogen) for expression. For the ZipACR constructs, the original EYFP fusion from Addgene (#88844) was retained and the construct was either inserted into the pcD-NA3.1+ vector or a second generation lentiviral vector pLenti with CMV promoter. No functional or fluorescent differences in the expressed protein were observed between expression from pcDNA3.1+ and pLenti vector in HEK 293 cells and the results were combined for comparison and analysis.

Site-directed mutagenesis of the ZipACR, Chrimson and GtACR2 were performed with overlap extension PCR using Phusion DNA Polymerase (Life Technologies) and ligated into the pcDNA3.1+ or pLenti expression vector.

For ChrimsonR and vfChrimson, additional improvements for membrane trafficking were used as previously described (Bonaventura et al. 2019). In brief, the N-terminal leading sequence from ChR1/oChIEF/ReaChR was transplanted to the transmembrane A-G of ChrimsonR and vfChrimson. A trafficking sequence was added after the transmembrane G before fusion to tdTomato. The modified ChrimsonR/vf-Chrimson-tdTomato constructs were inserted into the 2nd generation lentiviral pLenti vector with CMV promoter and used to make stable ChrimsonR/vfChrimson-tdTomato HEK 293 cell lines. The improved vf-Chrimson is named IvfChr.

For the comparison of IvfChr-citrine and Zip(151V)-Kv2.1-2A-IvfChr-citrine-Kv2.1 (ZipV-IvfChr) photocurrents, the two cDNAs were inserted into pcDNA3.1+ for expression. The IvfChr in both designs contains the modified N-terminal sequence from ChR1/oChIEF/ReaChR but the trafficking sequence was only used in the IvfChr-citrine design and not the Kv2.1 design.

For the expression of HcKCR1, HcKCR2 and WiChR, the mammalian codon-optimized sequences of the 3 potassium-selective ChRs were synthesized as gBlocks (IDT DNA) based on the published sequences and ligated into pcDNA3 or pcD-NA3.1H(+) vectors 5’ in-frame of coding sequences of EYFP or citrine using BamHI and NotI sites. Trafficking sequences were placed between potassium-selective ChRs and fluorescent protein. N-terminal signaling peptide from IvfChr was placed in some constructs tested. Membrane currents were similar in the presence and absence of N-terminal signaling peptides and the results were combined.

rAAV production

rAAV2-retro containing vfChrimson_EYFP_Kv2.1, IvfChr_ Citrine_Kv2.1, Zip(151T)-Kv2.1_IRES_DsRed, Zip(151T)-Kv2.1-2A-IvChrimson-citrine-Kv2.1 (ZipT-IvfChr), and Zip(151V)-Kv2.1-2A-IvfChr-citrine-Kv2.1 (ZipV-IvfChr) were produced according to the protocol at https://www.salk.edu/science/core-facilities/viral-vector-core/resources/. In brief, HEK 293A cells (Life Technologies, Carlsbad, CA) were grown to 90% confluence and transfected with vectors containing opsin proteins, and the helper plasmids XX6-80 and rAAV2-retro (gift from A. Y. Karpova, Howard Hughes Medical Institute Janelia Farm Research Campus). rAAV2-retro were released and purified according to protocol at https://www.takarabio.com/documents/User%20Manual/6232/6232_e.v1509Da.pdf.

Alternatively, virus vectors were produced at the University of Copenhagen, using triple-transfection protocol and purified with an iodixanol density gradient column following (Challis et al., 2019) protocol. In brief, 420 million HEK293t cells were seeded in DMEM 1965 (Glutamax, 10% FBS, 1% P/S). The day after, cells were transfected with 400 µg DNA in a 1:4:2 ratio (pAAV: rep/cap: pHelper), using linear PEI polyethylenimine (PEI, MW 25000). 18 hrs post-transfection, the media was exchanged to low serum DMEM 1965 (Glutamax, 1% FBS, 1% P/S) and incubated for 72 hrs before harvesting AAV vector from both media and cells using PEG8000. The AAVs were purified from an iodixanol density gradient column following 15 hrs ultracentrifugation (rotor SW28, 28000 rpm, 4 °C). Purity was confirmed with SDS-PAGE, and titer (vg/ mL) was determined with Quant-iT PicoGreen dsDNA Assay Kit (Invitrogen).

Electrophysiological characterization and imaging of HEK 293 cells

Whole-cell patch-clamping was performed with Axopatch 200A patch-clamp amplifier with Digidata 1322A data acquisition board and pCLAMP10 software on an Olympus IX73 inverted microscope. Cells were clamped at -60mV except in experiments with the ramp protocol and experiments comparing the 590 nm light-induced photocurrents between IvfChr-citrine and ZipV-IvfChr expressing cells. In the experiment comparing the IvfChr-citrine photocurrent to the 2A configuration, cells were clamped at -40mV to reduce the influence of ZipACR response with IvfChr photocurrent. Series resistance for the recorded cells were mostly under 15 MΩ and series resistance were compensated up to 70% during recordings where possible. For comparison of IvfChr and vfChrimson photocurrent in ZipV-IvfChr and BiPOLES-vfChrimson, the potassium gluconate based intracellular solution was used and cells were voltage-clamped at -40 mV (close to the reversal potential of gtACR2 and ZipACR) when stimulated with 590 nm light.

The extracellular solution was composed of (in mM): 145 NaCl, 3 KCl, 2 CaCl2, 1 MgCl2, 20 D-glucose and 10 HEPES (pH 7.35). For most experiments, the intracellular solution contained (in mM): 110 CsCl, 15 TEACl, 10 HEPES, 5 Cs2EGTA, 1 MgCl2 and 0.1mM CaCl2 (pH 7.25). In experiments comparing the 590 nm light-induced photocurrents between IvfChr-citrine and ZipV-IvfChr expressing cells and some ramp experiments, a K-gluconate based intracellular solution containing: 115 K-gluconate, 10 K2EGTA, 10 HEPES, 5 NaCl, 10 KCl, 2 MgATP and 0.1 Tris-GTP (pH 7.25), was used. For potassium-selective ChR recordings, only the K-gluconate-based intracellular solution was used.

For imaging the expression pattern of channelrhodopsin-expressing HEK 293 cells, cells on coverslips were imaged with an Olympus BX51 equipped with a 40x water immersion objective and a Hamamatsu Orca Flash 4.0LT sCMOS camera and X-cite LED110 light source. Images were acquired using Micromanager 1.4 with no gain and no electron multiplication and exposure time of 250 ms. Citrine and EYFP expression were imaged with a GFP filter cube (Semrock, FF02-470/30 excitation, FF01-520/35 emission and FF495-Di03 dichroic) and tdTomato was imaged with a TRITC filter cube (Semrock, FF01-543/22 excitation, FF01-593/40 emission and FF562-Di03 dichroic. For co-relating photocurrent amplitude and membrane expression, pre-recording images of the cells were captured with an Olympus IX73 microscope with a 40x air objective equipped with Photometrics Cascade II 512 EM-CCD camera and excitation by the Thorlabs 4 channel LED (LED4D118 with 405, 470, 530 and 590 nm channels and only 470 nm and 530 nm light were used for imaging with LED driving currents of 150 and 80 mA, respectively). An identical filter set was used as in the Olympus BX51 microscope. Images were acquired using Micromanager 1.4 with no gain and no electron multiplication and exposure time of 250 ms.

For most channelrhodopsin stimulation experiments 10 mW/ mm2 of 470 nm and 590 nm stimulation lights from Thorlabs 4 channel LED light source (LED4D118) was delivered to the specimen plane. The light intensity was controlled with a Thorlabs DC4104 4 channel controller electronically by the separate analogue outputs of a HEKA ITC-18 DAQ board under the control of WinWCP 5.5 software. In the spectral response experiment, a Sutter VF-5 system with a smart shutter in the neutral density mode was used to achieve cross spectrum stimulation at the same photon flux at 1.586 × 1015 photons / s / mm2 (25.1 mW/mm2 at 400 nm and 15.4 mW/mm2 at 650 nm) between 400-650 nm (25 nm increments). The VF-5 has the following filters and bandwidth settings: 451-398 nm with 15 nm bandwidth, 503-446 nm with 15 nm bandwidth, 564-498 nm with 14 nm bandwidth, 632-557 nm with 14 nm bandwidth, 703-627 nm with 13 nm bandwidth. A second Sutter shutter was placed in the light path before the light guide to control the light output. For experiments with ACRs, a 15 sec interval was used between stimulation episodes. For ChrimsonR/IvfChr stimulation, stimulation light pulses either have 15 sec intervals between stimulation episodes (no preconditioning) or a 15 sec interval – 1 sec 405 nm light pulse – 15 sec interval stimulation pattern for reconditioning of the protein with 405 nm before the subsequent stimulation. The stimulation light was guided into the fluorescent excitation light path of the Olympus IX73 microscope as with imaging but reflected with a Semrock FF685-DiO2 dichroic mirror to the objective.

Light intensity (in mW/mm2) was measured with a Thorlabs slide based light detector (S170C) at the specimen plane with illumination area defined by the microscope aperture and manually calculated for photon flux or illumination intensity.

For measuring the membrane expression of vfChrimson in various modified constructs, BiPOLES and 2A based cassettes, a membrane localized mCherry (encoded by pcDNA3-mCherry-CaaX) was co-expressed and live cell imaging were done on an Olympus Fluoview FV3000 confocal microscope with 40 × 0.8 NA water immersion objective. The citrine fluores-cence was acquired with a 488 nm laser excitation and 500-550 emission. mCherry images were acquired with a 594 nm laser excitation and 600-700 nm emission. For the analysis, single in-focus z-planes where the membrane boundaries were clearly defined were used. A straight line was drawn across the cell between 2 membrane boundaries devoid of the nucleus. The pixel intensities of the citrine and mCherry channels were measured to generate the Pearson Correlation coefficient.

Organotypic slices and virus infection

All procedures involving animals were approved by the Danish Animal Experiment Inspectorate. Hippocampi were isolated from postnatal P5-6 wild-type Sprague-Dawley rats (Janvier). The pups were decapitated and the brain removed and kept in an icy low sodium ACSF solution composed (in mM) of 1 CaCl2, 4 KCl, 1 MgCl2, 26 NaHCO3, 10 D-Glucose, 234 Sucrose, and 0,1% Phenol Red Solution, in milliQ H2O, and bubbled with a mixture of CO2 (5%) and O2 (95%). From that moment, all the steps were performed in a laminar flow tissue culture hood using sterile equipment and in total asepsis. The brain was poured into a petri dish filled with the low sodium ACSF solution for hippocampus dissection under microscope guidance (Stereomicroscope, Olympus x1 objective). After extraction, the hippocampi were sliced at 400 um thickness using an automatic tissue chopper (McILWAIN) and moved to a dish with pre-heated culture medium containing MEM Eagle medium 78.8% (Gibco), 20% heat-inactivated horse serum (Gibco), 1 mM L-glutamine, 1 mM CaCl2, 2mM MgSO4, 170 nM insulin, 0.0012% ascorbic acid, 12.9 mM D-glucose, 5.2 mM NaHCO3, 300 mM HEPES (Sigma), pH = 7.2-7.3, osmolarity adjusted to 315–325. The slices with intact DG and CA regions were then transferred onto air-fluid interface-style Milli-cell culture inserts (Millipore) in 6-well culture plates (ThermoFisherScientific) with 800 uL of sterile medium added below each insert. The slices were kept in a sterile incubator at 37 oC degrees with 5% CO2 (ThermoScientific, Steri-cycle i-160). The medium was replaced by a pre-warmed medium (37 oC) three times a week.

After two to three days of culture, the slices were microinjected in DG (or in CA1 when specified) along the characteristic horseshoe pattern with a pulled glass pipette containing the following viruses: retro-IvfChr (titer: 2.5×1012 particles/mL), retro-Zip(151T) (titer: 9×1012 particles/mL), retro-ZipT-IvfChr (titer: 2.9×1012 particles/mL), or retro-ZipV-IvfChr (titer: 1.5×1012 particles/mL). The injections were done under a microscope (using a Picospitzer III (Parker) connected to a Pulse Pal (10 ms pulse, every 0.5-sec). In the process of evaluating the R-ChR expression system in vivo, we noticed variability in the performance between batches from different virus sources. Although we adjusted for the nominal titers given by the providers, some batches produced toxicity while others fared well (data not shown). Therefore, the titers should not be taken as face value, and each batch must be tested in the region of interest before proceeding with the experiment. This concern applicable to any viral expression system has an added complication here. Higher titers, in addition to toxicity, may lead to the leakage of the soma-targeted channels expression in the terminals, which may cause unintended consequences. Recent developments in engineering soma-targeted motifs and peptides must be considered in future works (“Precision Calcium Imaging of Dense Neural Populations via a Cell-Body-Targeted Calcium Indicator” 2020) (Shemesh et al. 2020).

Whole-cell electrophysiology and light delivery in vitro

Two to three weeks post-infection, the organotypic slices were transferred into the recording chamber and continuously perfused with Artificial Cerebrospinal Fluid containing (in mM): 119 NaCL, 2.5 KCl, 26 NaHCO3, 1 NaH2PO4, supplemented with 11 D-glucose, 4 CaCl2, 4 MgCl2. Additionally, APV (50µM), NBQX (10µM) and picrotoxin (100µM) were added to block excitatory and inhibitory fast transmission. The solution was adjusted to pH 7.4 (osmolarity ∼330 mOsm) and bubbled with a mixture of CO2 (5%) and O2 (95%), at room temperature. The chamber was mounted on an upright microscope (Scientifica) linked to a digital camera (QImaging Exi Aqua) and the cells were visualized using 60X water-immersion objective (Olympus, LumiPlan). Acquisitions were performed in whole-cell configuration using Clampex 10.6 connected to a Multiclamp 700B amplifier via a Digidata 1550A digitizer (all from Molecular Devices). Voltage-clamp data were low-pass filtered at 200 Hz and digitized at 10 kHz and the whole-cell capacitance was compensated. Patch pipettes (2-4 MΩ of resistance) were filled with an internal solution containing (in mM): 153 K-gluconate, 10 HEPES, 4.5 NaCl, 9 KCl, 0.6 EGTA, 2 MgATP, and 0.3 NaGTP. The pH and osmolarity of the internal solution were close to physiological conditions (pH 7.4, osmolarity 297 mOsm). The access resistance of the cells in our sample was ∼ 25 MOhm. For voltage-clamp experiments, the Dentate Gyrus granule cells (otherwise specified when done in CA1 pyramidal cells) were clamped at -60mV. For current-clamp experiments, the cells were maintained at -60mV (except for the protocol where the potential was brought from -70 to -40mV) by constant current injection (holding current ± 100pA).

The recorded neurons were illuminated with 470 nm and 635 nm (nominal wavelengths) light with a coolLED pE-4000 system connected to the Digidata via a single TTL. The light irradiance was controlled via the control pod and was adjusted to 4mW/ mm2 for 470 nm and 7 mW/mm2 for 635 nm (otherwise specified). Light irradiance was measured with a Thorlabs digital optical power meter (PM100D) at the specimen plane with illumination area defined by the microscope aperture. Between the recordings, wavelengths were switched from 470 nm to 635 nm and vice versa. At the end of the recordings, the location of the recorded cell was confirmed by inspection at 10x magnification. Data analysis was performed using Clampfit 10.6 (Molecular device).

After recordings, slices were kept in 4% formalin, and later mounted under coverslips with Fluoromount (Sigma-Aldrich). Pictures were taken with an Apotome.2 (Zeiss) and ZEN software at 10X and 40X magnifications.

Acute slices

Experimental procedures were approved by the Animal Care and Use Committee of the University of Buenos Aires (CIC-UAL). Briefly, 4-week-old C57 mice (n = 4) were co-injected with AAV-8/2-hSyn1-ChrimsonVF_EYFP_Kv2.1 and AAV-8/2-hSyn1-ZipACR_Kv2.1_IRES_DsRed2 within the dentate gyrus of the hippocampus (250nl of a 1:1 mixture, coordinates from Bregma: 1 mm mediolateral, -2 mm anteroposterior, -1.5 dorsoventral). After 3 weeks of expression, the animals were sacrificed and the brain removed and cut in a solution composed of (in mM): 234 sucrose, 11 glucose, 26 NaHCO3, 2.5 KCl, 1.25 NaH2PO4, 10 MgSO4, and 0.5 CaCl2 (equilibrated with 95% O2–5% CO2). The slices were maintained at room temperature before being transferred in a recording chamber mounted on a microscope (Nikon) connected to a Mightex Illumination system for 470 nm and 635 nm light delivery (at 10mW/mm2). For current clamp experiments the amount of current injected was corrected in the inter-sweep interval to keep it close to -60 mV. In voltage clamp experiments holding potential was -60 mV.

Recordings were done in a solution of composition (in mM): 126 NaCl, 26 NaHCO3, 2.5 KCl, 1.25 NaH2PO4, 2 MgSO4, 2 CaCl2 and 10 Glucose (pH 7.4).

Patch-pipettes (2-4 MΩ of resistance) were filled with a K-glu-conate based internal solution (in mM, 130 K-gluconate, 5 KCl, 10 HEPES, 0.6 EGTA, 2.5 MgCl2·6H2O, 10 Phospho-creatine, 4 ATP-Mg, 0.4 GTP-Na3).

In vivo expression of ZipT-IvfChr and IvfChr in the facial nucleus

For expression of ZipT-IvfChr and IvfChr in the facial motor nucleus of the brainstem (FN), 8 to 10 weeks old c57Bl6/j mice (males) from Janvier (France) were anesthetized with Isoflurane (5% for induction, 2% for maintenance) and head-fixed on a stereotaxic frame (Model 940, Kopf, California) on top of a warm plate to keep their body-tem-perature at 37 oC. They were administered carprofen (5 mg/ kg, s.c.), then the skin was retracted, and a small craniotomy was drilled above the injection site. Suspension of either ret-roAAV-hSyn-ZipT-IvfChr (titer: 2.9×1012 particles/mL) or ret-roAAV-hSyn-IvfChr (titer: 2.5×1012 particles/mL) was injected unilaterally at a volume of 400nL and a rate of 2 nL per second, using a nanoliter injector (Nanoject III, Drummond, Broomal, PA). The coordinates for injection in the FN were -5.6 mm an-teroposterior, -1.4 mm mediolateral and -6.0 mm ventral, relative to bregma. After virus injection, the injection pipette was kept in place for an additional 10 minutes before being retracted. After they recovered from the surgery on a warm-plate, the animals were put back in their home-cages. They were housed in groups of 4 individuals per cage in a normal 12 h light/ dark cycle with food and water ad-libitum. Four to six weeks after the surgery, the animals were anesthetized with Urethane (Sigma-Aldrich, 1.5mg/kg for first injection, followed by injection of 1/5 of the initial dose until the reflex disappeared) and head-fixed on a KOPF stereotaxic frame (Model 940, Kopf, California). The reflexes were monitored during the entire procedure. The wound on the skin was reopened and the craniotomy above the injection site was re-open when necessary.

For optical stimulation of the FN motoneurons, a 200 um diameter optic-fiber connected to a laser source (Doric Lenses, Québec, Canada) was implanted above the facial nucleus at coordinates -5.6 mm anteroposterior, -1.4 mm mediolateral relative to bregma, and -4.1 to -4.5 mm ventral to the brain surface. Pulses of 100 ms of either 450 nm or 638 nm illumination at 20 mW/ mm2 were given at 0.33 Hz through a PulsePal trigger generator (Open Ephys) connected to the laser. The whiskers’ movements were recorded at 750 FPS with an Optronis Camera (CR3000, Optronis GMBH, Germany) and acquired through TimeBench (2.6.30, Optronis). After acquisition, the videos were exported as AVI with a 25 AVI framerate.

At the end of the experiment, the animals were sacrificed and the brain extracted and kept in Formalin 10% until slicing. 100 um sections were sliced with a Vibratom (Leica, VT1200S) and counterstained with DAPI (Thermofisher, D1306) and eGFP (Thermofisher, CAB4211) for histological localization of the virus expression, and imaged with a ZEISS Apotome microscope (Axio Imager M2). The experimenters were not blind to the group’s status; however, the automated analysis did not required blindness as this would not affect the outcome of the experiment

Whiskers tracking and analysis

The whiskers were tracked offline with a custom MatLab tool-box (available on GitHub GitHub - NabaviLab-Git/Whis-kerTrack: WhiskerTrack is a MATLAB toolbox to track the whiskers automatically with high accuracy.). In brief, 5 to 10 whiskers per animal were tracked frame by frame by using an automatic search method to fit the best line to the linear part of each whisker. The slope of this line was considered as the angle of its respective whisker. The absolute changes in the angle of each whisker were compared to its baseline and calculated as a Z-Score based on the mean and standard deviation.

The area under curve (AUC) was calculated as AUC= Z × W with Z being the averaged Z-Score for the maximum duration of the response and W being the window of the maximum response (200 ms from the onset of the light pulse).

For comparison between ZipT-IvfChr and IvfChr, we averaged the AUC of the responses to 450 nm or 635nm illumination over 3 trials and calculated the ratio of the AUC to 450 nm over 635nm. Statistical significance was calculated with a parametric t-test with MatLab.

Data Analysis

The data were analyzed and plotted using GraphPad Prism V9 (GraphPad Software, La Jolla, CA, USA). All values are indicated as mean ± standard deviation (SD) or mean ± standard error of the mean (SEM), as specified. For statistical comparison between groups, data were tested for normal distribution using D’Agostino and Pearson, Shapiro-Wilk or KS normality test. Comparison of the mean between groups has been done with a one-way analysis of variance (ANOVA) followed by Dunnett’s multiple comparisons test.

ZEN software (Zeiss) and FIJI (from ImageJ) were used for images processing.

Figure 1A, 1B1, 3E1, 4D and 5B1 have been created with BioRender.com.